---
title: "S3 vs R6 in rtfreporter — a hands-on tour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S3 vs R6 in rtfreporter — a hands-on tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
```

`rtfreporter` is designed to be a small but realistic case study in how to
pick between R's two main object systems, **S3** and **R6**, inside a
single package.  This vignette walks through the choices using the
package's own code as the example, and shows how to *see* the difference
in behaviour rather than just argue about it on paper.

```{r setup}
library(rtfreporter)
```

------------------------------------------------------------------------

## In one sentence

> **Default to S3.  Reach for R6 only when reference semantics genuinely
> change what the API can do.**

`rtfreporter` follows that rule:

| Class                                  | System                | Why                                                    |
|---------------------------------------|-----------------------|--------------------------------------------------------|
| `rtf_border_side`, `rtf_border`, `rtf_table_border` | **S3**         | Value objects.  Copy-on-modify is the right semantics. |
| `rtf_table_style`                     | **S3**                | Snapshot template; works fine without references.      |
| `rtftable`, `rtfplot`, `rtfreport`, `rtf_document`  | **S3**         | Data records; round-trip through `saveRDS()`.          |
| **`rtf_theme`**                       | **R6** *(optional)*   | Shared *mutable* defaults.  This is the one case S3 cannot match cleanly. |

Everything else — borders, the snapshot style, the pipe document, even
the internal renderer scaffold — is plain S3.  `Imports:` is empty;
R6 lives in `Suggests:` and is only needed when you opt in to
`rtf_theme()`.

------------------------------------------------------------------------

## 1. S3 in this package

S3 is R's lightweight class system: a value is "of class X" if its
`class` attribute contains `"X"`.  There is no formal type declaration,
no encapsulation, no inheritance enforcement.  The trade-off is
simplicity — and that simplicity is exactly what most rtfreporter
objects want.

### Tagged-list constructors

```{r}
b <- rtf_border(top = rtf_border_side(), bottom = rtf_border_side("double", 20L))
b
class(b)
str(b, max.level = 2)
```

`rtf_border()` is just a function that builds a named list and stamps
a class attribute on it.  You can poke around with `str()`, `dput()`
and `saveRDS()` — every field is plain data.

### S3 method dispatch

The `print()` you saw above is dispatched through `UseMethod()`.  The
package registers a small set of S3 methods (NAMESPACE):

```{r}
methods(class = "rtf_border")
methods(class = "rtf_table_border")
methods(class = "rtftable")
```

The `plot()` methods (covered in §3 below) are S3 too.  Adding a
new behaviour for `rtf_border` is as easy as defining a new
`fn.rtf_border` function and registering it.

### Copy semantics: borders are values, not handles

Because S3 lives in normal R values, *every* assignment makes a copy:

```{r}
b1 <- rtf_border_top()
b2 <- b1                 # copy, not a reference
b2$top <- NULL           # mutating b2 …
b1$top                   # … does not affect b1
```

That is exactly what you want for things like borders: they are
self-contained values.  Two tables that share the same border spec
cannot accidentally clobber each other.

### Functional derivation

When you do want a slightly modified copy, the package provides
explicit helper functions:

```{r}
heavy_top <- rtf_border_with(b1, top = rtf_border_side("thick", 30L))
heavy_top
identical(b1, heavy_top)   # b1 is unchanged
```

```{r}
tfl       <- rtf_table_style_tfl()
tfl_bold  <- rtf_table_style_with(tfl, header_bold = TRUE)
c(tfl$header_bold, tfl_bold$header_bold)
```

The pattern is "build, derive, build, derive" — fully functional,
nothing hidden behind references.

### The snapshot style: predictable, surprising-free

`rtf_table_style` is an S3 list that holds default values for
`rtftable()` to consume.  Critically, the values are **snapshotted
at the time `rtftable()` runs**:

```{r}
style <- rtf_table_style_tfl()
t1    <- rtftable(data.frame(A = 1:2), style = style)
style$header_bold <- TRUE                          # too late
t2    <- rtftable(data.frame(A = 1:2), style = style)
c(t1 = t1$col_spec[[1L]]$header_bold,
  t2 = t2$col_spec[[1L]]$header_bold)
```

`t1` snapshotted `header_bold = FALSE` at construction; the later
mutation to `style` only affects newly constructed tables.  This
behaviour is intuitive, easy to reason about, and the right default
for a defaults bag.

------------------------------------------------------------------------

## 2. The one R6 class: `rtf_theme`

There is exactly one situation where rtfreporter wants the
*opposite* of snapshot semantics: a **shared mutable theme** that
multiple tables observe in real time.  Suppose you have 20 tables in
a report, all using the same "company TFL look", and partway through
review you decide to flip `header_bold` on globally.

With the S3 style you would either rebuild all 20 tables or pre-empt
the change at construction.  With the R6 theme it is one assignment:

```{r, eval = requireNamespace("R6", quietly = TRUE)}
theme <- rtf_theme(header_bold = FALSE)
df1   <- data.frame(A = 1:2, B = c("x", "y"))
df2   <- data.frame(A = 3:4, B = c("p", "q"))
t1    <- rtftable(df1, theme = theme)
t2    <- rtftable(df2, theme = theme)

inherits(theme, "R6")
inherits(theme, "rtf_theme")

# Mutate the theme — every referencing table now renders bold headers
theme$header_bold <- TRUE
```

The trick is that `rtftable()` stores the R6 **reference** in
`tbl$theme` and the renderer calls a small internal helper
(`.refresh_theme(tbl)`) before each render to re-snapshot the
theme's current state.  So `t1` and `t2` reflect the mutation on
their very next render with no rebuild.

This is the one trick S3 cannot do cleanly: S3 lists are
copy-on-modify, so a mutation to `theme` would never propagate to
already-constructed tables.

### Why R6 is *Suggests*, not *Imports*

`rtf_theme()` is **gated** on the R6 package being installed.  All
other features — including the snapshot `rtf_table_style` — work
without R6.  Users who want shared mutable themes opt in with
`install.packages("R6")`; everyone else pays nothing.

```{r, eval = !requireNamespace("R6", quietly = TRUE)}
# Without R6 installed, rtf_theme() errors helpfully:
tryCatch(rtf_theme(), error = function(e) conditionMessage(e))
```

### When to choose `rtf_theme` over `rtf_table_style`

A small decision tree:

* **One table, one set of defaults?** Pass arguments directly to
  `rtftable()`.  No style, no theme.
* **Many tables, defaults locked at build time?** `rtf_table_style()`.
* **Many tables, defaults you might want to retune after building?**
  `rtf_theme()` — but you take on a *Suggests* dependency.

------------------------------------------------------------------------

## 3. S3 method dispatch in action: the `plot()` methods

Beyond `print()`, the package exposes a small `plot()` family that
takes advantage of S3 dispatch to give you a fast pre-render
inspection without leaving R.

### A border

```{r}
plot(rtf_border(top    = rtf_border_side("thick", 30L),
                bottom = rtf_border_side("double", 20L),
                left   = rtf_border_side("dot")))
```

### A table border (zones)

```{r}
plot(rtf_border_tfl())
```

### A table — layout wireframe

```{r}
df <- data.frame(USUBJID = c("001-001", "001-002", "001-003"),
                 TRT     = c("Placebo", "Active",  "Active"),
                 AVAL    = c(12.3, 14.1, 11.7))
tbl <- rtftable(df, col_rel_width = c(2, 1, 1), row_height_twips = 280L)
plot(tbl)
```

### A document — page thumbnails

```{r}
doc <- rtf_document()
doc <- rtf_section(doc, page = 1,
                   secinfo = list(header = rtf_header(c(l = "Demo")),
                                  footer = rtf_footer(c(c = "Confidential"))))
doc <- rtf_tables(doc, list(df, df, df))
plot(doc)
```

Each `plot.*()` method is one function.  Dispatch is handled by
`UseMethod()` — the same mechanism as `print()` and `summary()`.
That is the entire S3 method system: no class generators, no
formal interface declarations, no compile step.

------------------------------------------------------------------------

## 4. Putting the choice into your own packages

Use this list as a starting checklist.  Pick S3 unless several rows
push you towards R6.

| Indicator                                                       | Lean toward |
|-----------------------------------------------------------------|-------------|
| The object is a value the caller may freely copy.               | S3          |
| Round-tripping via `saveRDS()` / `dput()` matters.              | S3          |
| You want users to compose with `%>%` / `|>` (fresh copy each step). | S3      |
| The only methods you need are `print()` / `format()` / `summary()`. | S3      |
| Multiple holders must observe in-place mutations.               | **R6**      |
| The object has long-lived state that outlives a single call.    | **R6**      |
| You need a chainable mutating builder (`x$set_a()$set_b()`).    | R6 *(rare)* |
| You're tempted by R6 mostly for "everything is a method".        | Resist — that's a syntactic preference, not a semantic need. |

R6 is a powerful tool.  But every R6 class in your package adds a
dependency, hides behaviour behind methods, and breaks the simple
"R-values are immutable" mental model.  Used sparingly — exactly
once in rtfreporter — it gives you something you genuinely cannot
build cleanly any other way.  Used reflexively it just adds noise.

------------------------------------------------------------------------

## See also

* `LEARNING.md` at the repository root — the design log this
  vignette distils.
* `vignette("rtfreporter-quickstart")` — practical use of the
  package, not the design.
* `?rtf_theme` — full reference for the optional R6 theme.
