---
title: "gt -> rtfreporter: building RTF tables from gt objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gt -> rtfreporter: building RTF tables from gt objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
gt_ok <- requireNamespace("gt", quietly = TRUE)
knitr::opts_chunk$set(eval = gt_ok)
```

```{r setup, eval = TRUE}
library(rtfreporter)
```

```{r, include = FALSE, eval = !gt_ok}
cat("This vignette renders only when the `gt` package is installed.\n")
```

The [gt](https://gt.rstudio.com) package by Posit gives users a
friendly grammar for *display tables* -- `cols_label()`,
`tab_header()`, `tab_spanner()`, `cols_align()`,
`tab_source_note()`, `tab_footnote()`, ...  When you switch from the
HTML world to the clinical TFL world, you generally need the same
metadata to flow into the RTF document -- not be re-stated in a
different vocabulary.

Since v0.0.38, rtfreporter accepts a `gt_tbl` directly and can
optionally read any combination of nine GT attributes through to the
matching rtfreporter slot.  This vignette walks through the
end-to-end workflow.

## Quick start

```{r}
library(gt)

g <- gt(head(mtcars, 3)) |>
  cols_label(mpg = "MPG", cyl = "Cyl", disp = "Disp", hp = "HP",
             drat = "Drat", wt = "WT", qsec = "QSec",
             vs = "VS", am = "AM", gear = "Gear", carb = "Carb") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Demographics and baseline characteristics"
  ) |>
  tab_source_note("Source: built-in mtcars dataset.")
```

The simplest usage -- `read_gt = TRUE` pulls every supported
attribute through:

```{r}
doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(list(g), read_gt = TRUE)
```

Or use the standalone wrapper, which returns a plain `rtftable`:

```{r}
tbl <- as_rtftable(g, read = TRUE)
class(tbl)
```

## What gets read

`read_gt = TRUE` is shorthand for nine independent tokens, grouped
into three phases:

* **Phase A (shape-preserving, "safe")**
    * `"col_header"`   -- `cols_label()` -> the rtftable's column header
    * `"alignment"`    -- `cols_align()` -> per-column `col_spec[[j]]$align`
    * `"titles"`       -- `tab_header(title, subtitle)` -> the page
                          `titles[[i]]` block
    * `"source_notes"` -- `tab_source_note()` -> the page
                          `footnotes[[i]]` block

* **Phase B (structural, modifies the column space)**
    * `"spanning"` -- `tab_spanner()` -> multi-row col_header with
                      spanning cells
    * `"widths"`   -- `cols_width()` -> `column_widths_twips`
                      (`"NNpx"`) or `col_rel_width` (`"NN%"`)
    * `"hidden"`   -- `cols_hide()` -> drop the column before
                      rendering

* **Phase C (structural, modifies the row space + extra notes)**
    * `"footnotes"` -- `tab_footnote()` (any anchor) -> appended to
                       the page `footnotes[[i]]` block
    * `"stub"`      -- `groupname_col = ...`, `rowname_col = ...`,
                       `tab_stubhead()` -> group-transition rows and
                       a stubhead-labelled stub column

`read_gt = c("col_header", "titles")` opts in selectively.  Anything
listed there is read; anything left out falls back to the
rtfreporter default (e.g. column names from `names(data)`).

## End-to-end example: a TFL-style adverse-events table

This is the kind of table you would normally build by hand:

```{r}
ae <- data.frame(
  SOC = c("Nervous system disorders",
          "Nervous system disorders",
          "Nervous system disorders",
          "Gastrointestinal disorders",
          "Gastrointestinal disorders"),
  PT  = c("Headache", "Dizziness", "Insomnia",
          "Nausea",   "Diarrhoea"),
  Placebo_n = c("10 (33.3)", "2 ( 6.7)", "3 (10.0)",
                "6 (20.0)",  "4 (13.3)"),
  Active_n  = c("13 (43.3)", "2 ( 6.7)", "4 (13.3)",
                "7 (23.3)",  "5 (16.7)"),
  Total_n   = c("23 (38.3)", "4 ( 6.7)", "7 (11.7)",
                "13 (21.7)", "9 (15.0)"),
  stringsAsFactors = FALSE
)

ae_gt <-
  gt(ae, groupname_col = "SOC", rowname_col = "PT") |>
  cols_label(PT = "Preferred Term",
             Placebo_n = "Placebo",
             Active_n  = "Active",
             Total_n   = "Total") |>
  tab_stubhead(label = "SOC / PT") |>
  cols_align("right", columns = c(Placebo_n, Active_n, Total_n)) |>
  tab_spanner(label = "Treatment Arm",
              columns = c(Placebo_n, Active_n, Total_n)) |>
  tab_header(title    = "Table 14.2.1",
             subtitle = "Adverse Events Summary (Safety Population)") |>
  tab_source_note("Source: ADaM ADAE.") |>
  tab_footnote("Percentages are based on column N.")
```

Pipe it through `rtf_tables(read_gt = TRUE)`:

```{r}
doc <- rtf_document() |>
  rtf_section(
    page = 1,
    secinfo = list(
      header = rtf_header(rows = list(c(l = "Protocol RTF-101",
                                         r = "Page {AUTO_PAGE}"))),
      footer = NULL
    )
  ) |>
  rtf_tables(list(ae_gt), read_gt = TRUE)
```

`doc$contents[[1L]]` now has:

```{r}
tbl <- doc$contents[[1L]]
nrow(tbl$data)          # 5 PT rows + 2 SOC group rows
tbl$data$PT[1:3]        # SOC row + first two PTs (PT 1 is the group label)
tbl$col_header          # spanner row + stubhead-labelled bottom row
```

Page-level blocks:

```{r}
doc$titles[[1L]]        # title + subtitle
doc$footnotes[[1L]]     # footnote text + source note (in that order)
```

Render to RTF:

```{r}
out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

## Precedence: explicit arguments beat gt-extracted values

The translation honours the documented precedence chain.  If you
specify `col_header = c("A", "B", ...)` explicitly, gt's
`cols_label()` is ignored for that table:

```{r}
g <- gt(head(mtcars, 1)[, "mpg", drop = FALSE]) |>
  cols_label(mpg = "MPG")
tbl <- as_rtftable(g, read = TRUE, col_header = "Override")
tbl$col_header[[1L]]
```

The same applies to:

| Slot                  | Explicit arg                 |
|-----------------------|------------------------------|
| Column header         | `col_header`                 |
| Per-column alignment  | `col_spec[[j]]$align`        |
| Spanner stack         | (passed via `col_header`)    |
| Absolute widths       | `column_widths_twips`        |
| Relative widths       | `col_rel_width`              |
| Page titles           | `rtf_titles()`               |
| Page footnotes        | `rtf_footnotes()`            |

Per-column `col_spec` entries are **deep-merged** -- a user-supplied
`align = "left"` for column 1 only knocks out gt's column-1
alignment, the other columns keep gt's values.

## What is *not* yet supported

* Cell-mark injection.  GT footnotes can carry symbol marks (`¹`,
  `²`, ...) attached to specific cells.  rtfreporter v0.0.x emits the
  texts in the page footnote block but does not insert the marks
  into the body cells.
* `column_width` strings other than `"NNpx"` or `"NN%"`.  Mixed-unit
  widths bail out to `NULL` and the renderer falls back to its
  default proportional split.
* Cell-level styles (`tab_style()`).  Background colour, bold-cell
  patterns, etc. are still gt-only.
* Markdown rendering inside titles / labels.  `md("**bold**")` is
  flattened to the raw string `"**bold**"`; rtfreporter does not yet
  have a markdown subsystem.

## See also

* The full specification: `specs/gt-integration-spec.md` in the
  repository root.
* `?as_rtftable` and `?rtf_tables` for the per-argument docs.
* The `gt` package documentation: <https://gt.rstudio.com>.
