---
title: "rtfreporter Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtfreporter Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rtfreporter)
```

## Overview

`rtfreporter` is a clinical RTF reporting toolkit for tables, listings, and
figures. It is aimed at clinical-trial style outputs, not as a general-purpose
RTF library.

This quick start uses bundled test data to build two reports:

1. A one-page demographics report.
2. A laboratory report with one section per analyte and the analyte name in the section header.

## Test data

```{r data-load}
data_dir <- file.path("..", "tests", "testdata")
dm <- read.csv(file.path(data_dir, "dm.csv"), stringsAsFactors = FALSE)
ae <- read.csv(file.path(data_dir, "ae.csv"), stringsAsFactors = FALSE)
lb <- read.csv(file.path(data_dir, "lb.csv"), stringsAsFactors = FALSE)
head(dm)
head(ae)
head(lb)
```

## Common report styling

```{r common-style}
make_common_header <- function(study_title) {
  list(
    rows = list(
      c(l = "Protocol: RTF-101", r = "Page {PAGE} of {TOTAL_PAGES}"),
      c(l = study_title, r = "For Clinical Study Use Only")
    )
  )
}

make_common_footer <- function() {
  list(
    rows = list(
      c(l = "Confidential - Internal Use Only")
    )
  )
}
```

## Example 1: One-page DM report

```{r dm-report}
dm_report <- rtfreport$new()
dm_report$set_default_header(make_common_header("Table 14.1.1 Demographics"))
dm_report$set_default_footer(make_common_footer())

dm_sec <- dm_report$add_section(
  header = c(l = "Demographics (Screened Population)", r = "Safety Set")
)
dm_page <- dm_report$add_page(
  section_index = dm_sec,
  title = "Table 14.1.1 Demographics by Treatment Arm"
)

dm_disp <- data.frame(
  `Subject ID` = dm$USUBJID,
  `Treatment Arm` = dm$ARM,
  Sex = dm$SEX,
  Age = dm$AGE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dm_tbl <- rtftable$new(
  dm_disp,
  col_header = c("Subject ID", "Treatment Arm", "Sex", "Age"),
  col_rel_width = c(1.8, 2.2, 0.8, 0.8),
  row_height_twips = 300L,
  header_row_height_twips = 300L
)

dm_report$add_table(dm_sec, dm_page, data = dm_tbl, footer = "Source: DM domain")

dm_out <- file.path(tempdir(), "rtfreporter_quickstart_dm_report.rtf")
generate_rtfreport(dm_report, dm_out, overwrite = TRUE)
dm_out
```

The report uses a common protocol header, a section title, a page title, and a
clinical-style footer.

## Example 2: Laboratory report with one section per analyte

```{r lb-report}
lb_report <- rtfreport$new()
lb_report$set_default_header(make_common_header("Table 14.2.1 Clinical Laboratory Summary"))
lb_report$set_default_footer(make_common_footer())

lb_tests <- unique(lb$LBTEST)

for (test_name in lb_tests) {
  sec <- lb_report$add_section(
    header = c(l = paste0("Clinical Laboratory: ", test_name), r = "Page {PAGE} of {TOTAL_PAGES}")
  )
  page <- lb_report$add_page(
    section_index = sec,
    title = paste0(test_name, " by Subject")
  )

  test_df <- subset(lb, LBTEST == test_name, select = c(USUBJID, ARM, VISIT, LBSTRESN, LBSTRESU, LBNRLO, LBNRHI, LBNRIND))
  names(test_df) <- c("Subject ID", "Treatment Arm", "Visit", "Result", "Unit", "Ref Low", "Ref High", "Flag")

  test_tbl <- rtftable$new(
    test_df,
    col_rel_width = c(1.4, 1.6, 1.0, 0.8, 0.8, 0.8, 0.8, 0.7),
    row_height_twips = 300L,
    header_row_height_twips = 300L
  )

  lb_report$add_table(sec, page, data = test_tbl, footer = paste0("Source: LB domain | ", test_name))
}

lb_out <- file.path(tempdir(), "rtfreporter_quickstart_lab_report.rtf")
generate_rtfreport(lb_report, lb_out, overwrite = TRUE)
lb_out
```

Each section uses the analyte name in the header so the report reads naturally
as a clinical laboratory listing.

## Reusable section helper

When a section needs multiple pages from multiple data frames, the convenience
method `add_section_from_dataframes()` can create the section and append one
page per data frame in one call.

```{r helper-note}
report <- rtfreport$new()
report$add_section_from_dataframes(
  data_list = list("DM page" = dm[1:2, ], "AE page" = ae[1:2, ])
)
```

## Next steps

The package also provides `rtftable()` for table metadata, `rtfplot()` for
embedded figures, and `generate_rtfreport()` for writing the final document.
