---
title: "rtfreporter Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtfreporter Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rtfreporter)
```

## Overview

`rtfreporter` is a clinical RTF reporting toolkit for tables, listings, and
figures. It is aimed at clinical-trial style outputs, not as a general-purpose
RTF library.

This quick start uses bundled test data to build two reports:

1. A one-page demographics report.
2. A laboratory report with one section per analyte and the analyte name in the section header.

## Test data

```{r data-load}
data_dir <- file.path("..", "tests", "testdata")
dm <- read.csv(file.path(data_dir, "dm.csv"), stringsAsFactors = FALSE)
ae <- read.csv(file.path(data_dir, "ae.csv"), stringsAsFactors = FALSE)
lb <- read.csv(file.path(data_dir, "lb.csv"), stringsAsFactors = FALSE)
head(dm)
head(ae)
head(lb)
```

## Common report styling

Use `rtf_header()` and `rtf_footer()` to build reusable header/footer objects.
Both accept a single named vector (one row) or a `list` of named vectors
(multi-row). Keys `l`, `c`, `r` map to left, center, and right columns.

```{r common-style}
make_common_header <- function(study_title) {
  rtf_header(
    rows = list(
      c(l = "Protocol: RTF-101", r = "Page {AUTO_PAGE} of {TOTAL_PAGES}"),
      c(l = study_title,         r = "For Clinical Study Use Only")
    )
  )
}

make_common_footer <- function() {
  rtf_footer(c(l = "Confidential - Internal Use Only"))
}
```

## Example 1: One-page DM report

```{r dm-report}
dm_report <- rtfreport$new()

dm_sec <- dm_report$add_section(
  header = make_common_header("Table 14.1.1 Demographics"),
  footer = make_common_footer()
)
dm_page <- dm_report$add_page(
  section_index = dm_sec,
  title = "Table 14.1.1 Demographics by Treatment Arm"
)

dm_disp <- data.frame(
  `Subject ID` = dm$USUBJID,
  `Treatment Arm` = dm$ARM,
  Sex = dm$SEX,
  Age = dm$AGE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dm_tbl <- rtftable$new(
  dm_disp,
  col_header = c("Subject ID", "Treatment Arm", "Sex", "Age"),
  col_rel_width = c(1.8, 2.2, 0.8, 0.8),
  row_height_twips = 300L,
  header_row_height_twips = 300L
)

dm_report$add_table(dm_sec, dm_page, data = dm_tbl, footer = "Source: DM domain")

dm_out <- file.path(tempdir(), "rtfreporter_quickstart_dm_report.rtf")
generate_rtfreport(dm_report, dm_out, overwrite = TRUE)
dm_out
```

The report uses a common protocol header, a section title, a page title, and a
clinical-style footer.

## Example 2: Laboratory report with one section per analyte

```{r lb-report}
lb_report <- rtfreport$new()
lb_tests <- unique(lb$LBTEST)

for (test_name in lb_tests) {
  sec <- lb_report$add_section(
    header = rtf_header(
      rows = list(
        c(l = "Protocol: RTF-101",                    r = "Page {AUTO_PAGE} of {TOTAL_PAGES}"),
        c(l = paste0("Table 14.2.1 Clinical Lab: ", test_name), r = "For Clinical Study Use Only")
      )
    ),
    footer = make_common_footer()
  )
  page <- lb_report$add_page(
    section_index = sec,
    title = paste0(test_name, " by Subject")
  )

  test_df <- subset(lb, LBTEST == test_name,
                    select = c(USUBJID, ARM, VISIT, LBSTRESN, LBSTRESU, LBNRLO, LBNRHI, LBNRIND))
  names(test_df) <- c("Subject ID", "Treatment Arm", "Visit", "Result", "Unit",
                      "Ref Low", "Ref High", "Flag")

  test_tbl <- rtftable$new(
    test_df,
    col_rel_width = c(1.4, 1.6, 1.0, 0.8, 0.8, 0.8, 0.8, 0.7),
    row_height_twips = 300L,
    header_row_height_twips = 300L
  )

  lb_report$add_table(sec, page, data = test_tbl,
                       footer = paste0("Source: LB domain | ", test_name))
}

lb_out <- file.path(tempdir(), "rtfreporter_quickstart_lab_report.rtf")
generate_rtfreport(lb_report, lb_out, overwrite = TRUE)
lb_out
```

Each section specifies its header and footer directly. If a section omits
`header` or `footer`, the value is inherited from the immediately preceding
section.

## Reusable section helper

When a section needs multiple pages from multiple data frames, the convenience
method `add_section_from_dataframes()` creates the section and appends one
page per data frame in a single call.

```{r helper-note}
report <- rtfreport$new()
report$add_section_from_dataframes(
  data_list        = list("DM page" = dm[1:2, ], "AE page" = ae[1:2, ]),
  section_header   = make_common_header("Bundle Demo"),
  section_footer   = make_common_footer()
)
```

## Page tokens

Use these tokens in header/footer text:

| Token | Behavior |
|-------|----------|
| `{AUTO_PAGE}` | Dynamic page number updated by the RTF viewer (`\chpgn`) — **recommended** |
| `{PAGE}` | Alias for `{AUTO_PAGE}` |
| `{TOTAL_PAGES}` | Static total page count computed when `generate_rtfreport()` is called |
| `{AUTO_TOTAL_PAGES}` | Alias for `{TOTAL_PAGES}` |

Recommended usage: `"Page {AUTO_PAGE} of {TOTAL_PAGES}"`.

## Next steps

The package also provides `rtftable()` for table metadata, `rtfplot()` for
embedded figures, and `generate_rtfreport()` for writing the final document.
