---
title: "Pipe-Based API for rtfreporter"
author: "ichirio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipe-Based API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The rtfreporter package provides two complementary APIs for building RTF reports:

1. **S3 Methods API** (v0.0.6): Mutable object manipulation with `add_section()`, `add_page()`, etc.
2. **Tidyverse Pipe API** (v0.0.6+): Immutable functional composition with `%>%` pipes

This vignette demonstrates the **Tidyverse Pipe API**, which is inspired by ggplot2 and dplyr, providing a clean, composable interface for building complex RTF reports.

## Quick Start

The basic workflow is:

1. Initialize a document with `rtf_document()`
2. Configure document settings with `rtf_config()` (optional)
3. Add content (tables/figures) with `rtf_tables()` and `rtf_figures()`
4. Define sections with `rtf_section()`
5. Apply formatting with `rtf_table_format()`, `rtf_header_format()`, etc.
6. Generate RTF with `generate_rtfreport()`

```{r, eval = FALSE}
library(rtfreporter)

# Create a simple report
doc <- rtf_document() %>%
  rtf_tables(list(mtcars)) %>%
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL))

# Generate RTF file
generate_rtfreport(doc, "report.rtf")
```

## API Functions

### Initialization and Configuration

#### `rtf_document()`

Creates a new RTF document object with clinical trial defaults.

**Defaults:**
- Page: Landscape letter (11" × 8.5")
- Margins: 0.75" top/bottom, 0.5" left/right
- Font: Courier
- Colors: Black

```{r, eval = FALSE}
doc <- rtf_document()

# With custom defaults
doc <- rtf_document(
  font_table = list(list(name = "Arial")),
  page = list(
    orientation = "portrait",
    width_in = 8.5,
    height_in = 11
  )
)
```

#### `rtf_config()`

Update document-level settings. Typically called once after creating a document.

**Parameters:**
- `font_table`: Override font table
- `color_table`: Override color table
- `page`: Override page settings
- `default_format`: Override default formatting

All parameters are **optional** (NULL = no change), making it safe for repeated calls:

```{r, eval = FALSE}
doc <- rtf_document() %>%
  rtf_config(page = list(orientation = "portrait"))
```

### Content Addition

#### `rtf_tables()`

Add one or more data.frames as content pages. Pages are auto-numbered.

```{r, eval = FALSE}
# Single table per page
doc <- rtf_document() %>%
  rtf_tables(list(df1, df2, df3))

# Multiple tables on one page
doc <- rtf_document() %>%
  rtf_tables(list(df1, list(df2a, df2b), df3))
# Results in: df1 on page 1, df2a+df2b on page 2, df3 on page 3
```

#### `rtf_figures()`

Add image files as figure pages. Each image creates one new page.

```{r, eval = FALSE}
doc <- rtf_document() %>%
  rtf_figures(list("plot1.png", "plot2.png"))
```

### Structure Definition

#### `rtf_section()`

Map page numbers to sections with headers and footers.

```{r, eval = FALSE}
# Single section starting at page 1
doc <- rtf_document() %>%
  rtf_tables(list(df1, df2, df3)) %>%
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(l = "Analysis", r = "Table 1")),
    footer = rtf_footer(rows = list(c = "Page {CURRENT_PAGE}"))
  ))

# Multiple sections
doc <- rtf_document() %>%
  rtf_tables(list(df1, df2, df3)) %>%
  rtf_section(
    page = c(1, 3),
    secinfo = list(
      list(header = h1, footer = f1),  # Section 1 (pages 1-2)
      list(header = h2, footer = f2)   # Section 2 (pages 3+)
    )
  )
```

### Formatting

All format functions follow the same pattern:
- Accept `pages` parameter: `"all"` or `c(1, 3, 5)` for specific pages
- All other parameters are **optional** (NULL = no change)
- Safe for multiple calls - later calls override only the specified parameters

#### `rtf_table_format()`

Format table content across pages.

```{r, eval = FALSE}
doc <- rtf_document() %>%
  rtf_tables(list(df1, df2, df3)) %>%
  rtf_table_format(pages = "all", border = "tfl", row_height_twips = 280L) %>%
  rtf_table_format(pages = 1, border = "none")  # Override page 1
```

#### `rtf_header_format()` and `rtf_footer_format()`

Format table headers and footers.

```{r, eval = FALSE}
doc <- doc %>%
  rtf_header_format(pages = "all", border = "top", row_height_twips = 280L) %>%
  rtf_footer_format(pages = c(1, 3), border = "top")
```

#### `rtf_figure_format()`

Format figure content.

```{r, eval = FALSE}
doc <- doc %>%
  rtf_figure_format(pages = "all", width_twips = 8000L, height_twips = 6000L)
```

## Complete Example

```{r, eval = FALSE}
library(rtfreporter)
library(magrittr)  # For %>% pipe

# Prepare data
df_safety <- data.frame(
  Event = c("Headache", "Nausea", "Dizziness"),
  Mild = c(5, 3, 2),
  Moderate = c(2, 1, 1),
  Severe = c(0, 0, 1)
)

df_efficacy <- data.frame(
  Response = c("Complete", "Partial", "No Response"),
  Count = c(25, 15, 10),
  Percent = c("50%", "30%", "20%")
)

# Create report
report <- rtf_document() %>%
  # Configure document (one-time)
  rtf_config(page = list(
    orientation = "landscape",
    width_in = 11,
    height_in = 8.5
  )) %>%
  # Add content
  rtf_tables(list(df_safety, df_efficacy)) %>%
  # Define sections with headers/footers
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(
      l = "Clinical Study Report",
      r = "Safety Analysis"
    )),
    footer = rtf_footer(rows = list(
      c = "Page {CURRENT_PAGE} of {AUTO_TOTAL_PAGES}"
    ))
  )) %>%
  rtf_section(page = 2, secinfo = list(
    header = rtf_header(rows = list(
      l = "Clinical Study Report",
      r = "Efficacy Analysis"
    )),
    footer = rtf_footer(rows = list(
      c = "Page {CURRENT_PAGE} of {AUTO_TOTAL_PAGES}"
    ))
  )) %>%
  # Apply formatting (multiple calls safe, later ones override)
  rtf_table_format(pages = "all", border = "tfl", row_height_twips = 280L) %>%
  rtf_header_format(pages = "all", border = "top", row_height_twips = 300L)

# Generate RTF file
generate_rtfreport(report, "clinical_report.rtf", overwrite = TRUE)
```

## Pipe vs. Methods API

### Using Pipes (Recommended)

```{r, eval = FALSE}
report <- rtf_document() %>%
  rtf_tables(list(df1, df2)) %>%
  rtf_section(page = 1, secinfo = sec_info) %>%
  rtf_table_format(pages = "all", border = "tfl")
```

### Using Methods (v0.0.6)

```{r, eval = FALSE}
report <- rtfreport()
sec <- report$add_section(header = h1, footer = f1)
report$add_page(section_index = sec, content = list(rtftable(df1)))
report$add_page(section_index = sec, content = list(rtftable(df2)))
report$set_default_page(...)
```

Both APIs work and produce identical RTF output. Choose based on your preference:
- **Pipes**: Functional style, easier to compose, modern R idiom
- **Methods**: Direct object manipulation, more explicit control

## Safe Parameter Defaults

The pipe API uses **NULL-safe parameters** to prevent unintended overwrites:

```{r, eval = FALSE}
# This is safe - only updates page 1
doc <- doc %>%
  rtf_table_format(pages = 1, border = "none")

# Later call doesn't overwrite - NULL means "no change"
doc <- doc %>%
  rtf_table_format(pages = "all", row_height_twips = 300L)
# Result: page 1 has border="none", row_height=300L
#         other pages have border="tfl", row_height=300L
```

## Object Structure

The S3 `rtf_document` object contains:

```r
list(
  document = list(
    font_table = ...,
    color_table = ...,
    page = list(orientation, width_in, height_in, margins),
    default_format = ...
  ),
  contents = list(...),          # Pages (data.frames or file paths)
  sections = list(...),          # Page → section info mapping
  table_formats = list(...),     # Format specs by page
  header_formats = list(...),
  footer_formats = list(...),
  figure_formats = list(...)
)
```

Each function returns a modified copy, enabling immutable composition.

## See Also

- `vignette("rtfreporter-quickstart")` - S3 Methods API (v0.0.6)
- `help(rtf_document)` - Detailed function documentation
- `generate_rtfreport()` - Render document to RTF file
