---
title: "rtfreporter — Clinical RTF Reporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtfreporter — Clinical RTF Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      eval = FALSE)   # set TRUE locally to regenerate outputs
library(rtfreporter)
```

## Overview

`rtfreporter` produces clinical-trial–style RTF reports (tables, listings,
figures) without depending on r2rtf or any other RTF library.

**Document hierarchy:** Document → Section → Page → Block
Each section carries its own `header` / `footer`.  When you omit them, the
previous section's values are inherited automatically.

---

## 1  Minimal one-page report

```{r minimal}
report <- rtfreport()

sec <- add_section(report,
  header = rtf_header(rows = list(
    c(l = "Protocol: RTF-101",          r = "ACME Pharma"),
    c(l = "Table 14.1.1 Demographics",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}")
  )),
  footer = rtf_footer(rows = list(
    c(l = "Source: DM domain.", r = "CONFIDENTIAL")
  ))
)

df <- data.frame(
  Subject = c("001", "002", "003"),
  Arm     = c("Active", "Placebo", "Active"),
  Age     = c(34L, 45L, 28L),
  Sex     = c("M", "F", "M"),
  stringsAsFactors = FALSE
)

add_page(report,
  section_index = sec,
  content = list(
    rtftable(df,
      col_rel_width    = c(2, 2, 1, 1),
      row_height_twips = 280L)
  ),
  footer_notes = "Source: DM domain. One row per subject."
)

generate_rtfreport(report, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 2  Lab shift table — one section per analyte

The shift table is the canonical multi-section clinical TFL.  Each analyte
(ALT, AST, HGB, …) occupies its own RTF section so the **analyte name
appears in the section header bottom-left**, changing from page to page.

RTF headers and footers are per-section; this is the correct way to vary
header content across pages of the same document.

```{r shift-table}
library(rtfreporter)

# --- synthetic shift-count data ----------------------------------------------
make_shift_df <- function(a_low, a_norm, a_high, b_low, b_norm, b_high) {
  data.frame(
    Baseline = c("Low", "Normal", "High"),
    A_Low    = a_low,  A_Normal = a_norm,  A_High = a_high,
    B_Low    = b_low,  B_Normal = b_norm,  B_High = b_high,
    stringsAsFactors = FALSE, check.names = FALSE
  )
}

lab_data <- list(
  list(label = "ALT (Alanine Aminotransferase)",
       df    = make_shift_df(c(4,0,0), c(1,13,1), c(0,2,3),
                             c(3,0,0), c(1,14,0), c(0,1,4))),
  list(label = "AST (Aspartate Aminotransferase)",
       df    = make_shift_df(c(5,0,0), c(0,11,2), c(0,1,5),
                             c(4,0,0), c(1,12,1), c(0,2,4))),
  list(label = "HGB (Hemoglobin)",
       df    = make_shift_df(c(2,1,0), c(1,16,1), c(0,1,2),
                             c(3,0,0), c(0,15,2), c(0,2,2)))
)

# --- shared table structure --------------------------------------------------
spanning_hdr <- list(
  list(from = 2L, to = 4L, label = "Treatment A  (N=24)", underline = TRUE),
  list(from = 5L, to = 7L, label = "Treatment B  (N=24)", underline = TRUE)
)
col_hdr    <- c("Baseline", "Low", "Normal", "High", "Low", "Normal", "High")
col_widths <- c(2160L, 900L, 900L, 900L, 900L, 900L, 900L)

col_spec_shift <- lapply(seq_len(7L), function(j)
  list(col = j, align = if (j == 1L) "left" else "center"))

make_shift_table <- function(df) {
  rtftable(
    data                = df,
    col_header          = col_hdr,
    spanning_header     = spanning_hdr,
    col_spec            = col_spec_shift,
    column_widths_twips = col_widths,
    border              = "tfl",
    row_height_twips    = 280L
  )
}

# --- common footer (same for all sections) -----------------------------------
common_footer <- rtf_footer(
  rows = list(
    c(l = paste0("ALT=Alanine Aminotransferase; ",
                 "AST=Aspartate Aminotransferase; HGB=Hemoglobin")),
    c(l = "Low/Normal/High: site reference range categories.  n = number of subjects.",
      r = "CONFIDENTIAL")
  )
)

# --- build report: one section per analyte -----------------------------------
report <- rtfreport()

for (item in lab_data) {
  sec_header <- rtf_header(
    rows = list(
      c(l = "Protocol: LAB-001",                           r = "ACME Pharma"),
      c(l = "Study Title: Phase III Safety Lab Assessment", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
      c(c = "Table 14.3.1  Laboratory Shift Table (Safety Population)"),
      c(l = item$label)       # ← analyte name: bottom-left, unique per section
    )
  )

  sec <- add_section(report, header = sec_header, footer = common_footer)
  add_page(report,
    section_index = sec,
    content       = list(make_shift_table(item$df)),
    footer_notes  = paste0("Source: LB domain.  ",
      "Subjects with both a baseline and at least one post-baseline assessment included.")
  )
}

generate_rtfreport(report, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

The resulting file has **3 sections** (one per analyte), each with its own
header row 4 set to the analyte label.

---

## 3  Embedding figures (ggplot2)

`rtfplot` embeds a PNG or JPEG image binary directly into the RTF `\pict` block.
The default alignment is **center**; specify `width_twips` to control the
display size.

```{r figure}
library(ggplot2)

# --- build a ggplot and save to a temp PNG -----------------------------------
p <- ggplot(
  data.frame(x = 1:10, y = (1:10)^2 + rnorm(10, sd = 3)),
  aes(x = x, y = y)) +
  geom_point(size = 3, colour = "#2166AC") +
  geom_smooth(method = "lm", se = FALSE, colour = "#D6604D") +
  labs(title = "Lab value vs. Visit (simulated)", x = "Visit", y = "Lab value") +
  theme_bw(base_size = 11)

png_path <- tempfile(fileext = ".png")
ggsave(png_path, plot = p, width = 7, height = 4.5, dpi = 150)

# --- embed in RTF ------------------------------------------------------------
plot_obj <- rtfplot(
  path        = png_path,
  width_twips = 9000L   # ~6.25 in; height computed from aspect ratio
)

report <- rtfreport()
sec <- add_section(report, 
  header = rtf_header(rows = list(
    c(l = "Protocol: FIG-001",           r = "ACME Pharma"),
    c(l = "Figure 14.1.1  Lab Value vs. Visit (Safety Population)")
  )),
  footer = rtf_footer(rows = list(c(l = "Source: LB domain.")))
)
add_page(report, sec, content = list(plot_obj))

generate_rtfreport(report, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 4  Figure and table on the same page

Keep the figure width moderate so both elements fit on one page.

```{r fig-plus-table}
tbl_summary <- rtftable(
  data.frame(
    Statistic = c("n", "Mean", "SD"),
    Value     = c("30", "42.7", "8.4"),
    stringsAsFactors = FALSE
  ),
  table_width_pct  = 40,     # 40 % of writable width
  table_align      = "left",
  row_height_twips = 280L
)

report2 <- rtfreport()
sec2 <- add_section(report2, 
  header = rtf_header(rows = list(
    c(l = "Protocol: FIG-001", r = "ACME Pharma"),
    c(l = "Figure + Summary Table")
  )),
  footer = rtf_footer(rows = list(c(l = "Source: LB domain.", r = "CONFIDENTIAL")))
)
add_page(report2, 
  section_index = sec2,
  content = list(
    rtfplot(png_path, width_twips = 7200L),
    tbl_summary
  )
)
generate_rtfreport(report2, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 5  Table width and alignment

| Parameter | Description |
|---|---|
| `table_width_pct = 70` | 70 % of writable page width |
| `table_width_twips = 9000L` | Absolute: 9000 twips = 6.25 in |
| `table_align = "center"` | Center the table on the page |
| `table_align = "right"` | Right-align the table |

```{r table-sizing}
df <- data.frame(A = 1:3, B = letters[1:3], C = c(1.1, 2.2, 3.3))

# 70 % of writable width, left-aligned (default)
tbl_left <- rtftable(df, table_width_pct = 70)

# Absolute width, centered
tbl_center <- rtftable(df,
  table_width_twips = 9000L,
  table_align       = "center")

# Absolute width, right-aligned, with relative column widths 3:1:1
tbl_right <- rtftable(df,
  table_width_twips = 7200L,
  table_align       = "right",
  col_rel_width     = c(3, 1, 1))

# Full writable width (100 %)
tbl_full  <- rtftable(df, table_width_pct = 100)
```

---

## 6  Column width control

Three ways to set column widths, in order of precedence:

| Method | Parameter | Notes |
|---|---|---|
| Absolute | `column_widths_twips = c(2880L, 1440L, 1440L)` | One value per column |
| Relative | `col_rel_width = c(2, 1, 1)` + `table_width_twips` | Proportional split |
| Auto | `auto_col_widths(df)` | Font-aware estimation |

```{r col-widths}
# Absolute column widths (twips)
tbl_abs <- rtftable(df,
  column_widths_twips = c(4320L, 2160L, 2160L))

# Relative widths 2:1:1 within a 7-inch table
tbl_rel <- rtftable(df,
  col_rel_width     = c(2, 1, 1),
  table_width_twips = 10080L)

# Auto-sizing: widths computed from the widest cell in each column
w <- auto_col_widths(df, table_width_twips = 14400L)
tbl_auto <- rtftable(df, column_widths_twips = w)
```

---

## 7  Header / footer layout

Keys `l`, `c`, `r` map to left, center, right alignment.
An unnamed single element defaults to **center** (for both header and footer).

```{r hf-layout}
# Single row, two columns
sec_hf1 <- rtf_header(rows = list(
  c(l = "Protocol: RTF-101", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}")
))

# Multi-row header
sec_hf2 <- rtf_header(rows = list(
  c(l = "Protocol: RTF-101",    r = "ACME Pharma"),
  c(l = "Phase III Safety Study", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(c = "Table 14.3.1  Shift Table (Safety Population)"),
  c(l = "ALT (Alanine Aminotransferase)")   # bottom-left: analyte name
))

# Footer with two rows; top border on row 1 only (default)
sec_ftr <- rtf_footer(rows = list(
  c(l = "ALT=Alanine Aminotransferase",     r = "CONFIDENTIAL"),
  c(l = "Low/Normal/High: site reference range categories.")
))

# Single unnamed row → center (default)
ftr_center <- rtf_footer(rows = list(c("Confidential — Internal Use Only")))
```

---

## 8  Page tokens

| Token | Behaviour |
|---|---|
| `{AUTO_PAGE}` | Dynamic page number (`\chpgn`), updated per page by the viewer |
| `{AUTO_TOTAL_PAGES}` | Dynamic NUMPAGES field, updated by the viewer — **recommended for assembled docs** |
| `{TOTAL_PAGES}` | Static total computed at render time by `generate_rtfreport()` |
| `{SECTION_PAGES}` | Dynamic SECTIONPAGES field (pages in the current section) |

Recommended: `"Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"`.

Use `{TOTAL_PAGES}` only when generating a single, self-contained file where
the total page count is known at render time.

---

## 9  Spanning headers

```{r spanning}
df_shift <- data.frame(
  Baseline = c("Low", "Normal", "High"),
  A_Low = c(4L, 0L, 0L), A_Norm = c(1L, 13L, 1L), A_High = c(0L, 2L, 3L),
  B_Low = c(3L, 0L, 0L), B_Norm = c(1L, 14L, 0L), B_High = c(0L, 1L, 4L)
)

tbl_span <- rtftable(
  data = df_shift,
  col_header = c("Baseline", "Low", "Normal", "High", "Low", "Normal", "High"),
  spanning_header = list(
    list(from = 2L, to = 4L, label = "Treatment A  (N=24)", underline = TRUE),
    list(from = 5L, to = 7L, label = "Treatment B  (N=24)", underline = TRUE)
  ),
  column_widths_twips = c(2160L, 900L, 900L, 900L, 900L, 900L, 900L),
  col_spec = lapply(seq_len(7L), function(j)
    list(col = j, align = if (j == 1L) "left" else "center")),
  border           = "tfl",
  row_height_twips = 280L
)
```

---

## 10  Multi data.frame table

When `data` is a list of data.frames each is rendered in sequence, with its
own column-header rows.  This lets you combine cohorts or time points in one
table without duplicating the spanning header.

```{r multi-df}
df_a <- data.frame(Label = c("Low", "Normal", "High"), n = c(4L, 13L, 3L))
df_b <- data.frame(Label = c("Low", "Normal", "High"), n = c(3L, 14L, 4L))

tbl_multi <- rtftable(
  data       = list(df_a, df_b),
  col_header = list(
    c("Baseline", "Arm A"),   # header for df_a
    c("Baseline", "Arm B")    # header for df_b
  ),
  column_widths_twips = c(3600L, 2160L),
  border              = "tfl",
  row_height_twips    = 280L
)
```

---

## 11  Border customization

```{r borders}
# TFL clinical preset (default): header top+bottom, last-row bottom
tbl_tfl <- rtftable(df, border = "tfl")

# Custom per-zone
tbl_custom <- rtftable(df,
  border = rtf_table_border(
    header   = rtf_border(top = rtf_border_side(), bottom = rtf_border_side()),
    last_row = rtf_border(bottom = rtf_border_side("double", 10L))
  )
)

# Header with bottom divider, footer with thick blue top border
hdr_bordered <- rtf_header(
  rows   = list(c(c = "Study Title")),
  border = rtf_border(bottom = rtf_border_side("single"))
)
ftr_blue <- rtf_footer(
  rows   = list(c(l = "Confidential")),
  border = rtf_border_top(style = "thick", width = 20L, color = "#003366")
)
```

---

## 12  Assembling multiple RTF files

`assemble_rtf()` joins independently generated RTF files into one document.
Use dynamic tokens (`{AUTO_TOTAL_PAGES}`) so page numbers are correct across
the assembled output.

```{r assemble}
make_rtf <- function(df, hdr_text, path) {
  r <- rtfreport()
  sec <- add_section(r,
    header = c(l = hdr_text, r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    footer = c(l = "CONFIDENTIAL")
  )
  add_page(r, section_index = sec, content = list(df))
  generate_rtfreport(r, path, overwrite = TRUE)
}

f1 <- tempfile(fileext = ".rtf"); make_rtf(mtcars[1:3, ], "Table 14.1 DM", f1)
f2 <- tempfile(fileext = ".rtf"); make_rtf(mtcars[4:6, ], "Table 14.2 AE", f2)

assembled <- tempfile(fileext = ".rtf")
assemble_rtf(c(f1, f2), assembled, overwrite = TRUE)
```

---

## 13  Auto column widths

`auto_col_widths()` estimates widths from the actual content and scales them
to a target table width.

```{r auto-widths}
df_ae <- data.frame(
  USUBJID  = c("001-001", "001-002"),
  AETERM   = c("Headache", "Nausea"),
  AESTDTC  = c("2024-01-15", "2024-02-03"),
  AESER    = c("N", "N"),
  stringsAsFactors = FALSE
)

# Scale to full writable width (14400 twips = 10 in landscape letter)
w <- auto_col_widths(df_ae,
  col_header        = "Subject ID | AE Term | Start Date | Serious",
  table_width_twips = 14400L)

tbl_ae <- rtftable(df_ae,
  col_header          = c("Subject ID", "AE Term", "Start Date", "Serious"),
  column_widths_twips = w,
  border              = "tfl",
  row_height_twips    = 280L)
```

---

## Quick-reference

### rtftable parameters

| Parameter | Type | Purpose |
|---|---|---|
| `data` | data.frame or list | Content rows |
| `col_header` | chr / list | Column header text |
| `spanning_header` | list | Spanning (grouped) header |
| `col_spec` | list | Per-column align / bold / italic / indent |
| `border` | `"tfl"` / object | Cell borders |
| `col_rel_width` | numeric vector | Proportional column widths |
| `column_widths_twips` | integer vector | Absolute column widths |
| `table_width_twips` | integer | Table total width (twips) |
| `table_width_pct` | numeric 0–100 | Table width as % of page |
| `table_align` | `"left"` / `"center"` / `"right"` | Table position on page |
| `row_height_twips` | integer | Data row height |
| `blank_rows` | integer vector | Insert blank separator rows |

### rtfplot parameters

| Parameter | Default | Purpose |
|---|---|---|
| `path` | — | PNG or JPEG file path |
| `width_twips` | `NULL` (full width) | Display width |
| `height_twips` | `NULL` (aspect ratio) | Display height |
| `align` | `"center"` | `"left"` / `"center"` / `"right"` |

### Header / footer single-row defaults

| Keys used | Columns | Alignments |
|---|---|---|
| `l=` only | 1 | left |
| `r=` only | 1 | right |
| `c=` only | 1 | center |
| `l=` + `r=` | 2 | left, right |
| unnamed single value | 1 | **center** (default) |
| `l=` + `c=` + `r=` | 3 | left, center, right |
