---
title: "Headers and footers (sections)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

A rtfreporter document is a flat sequence of content pages (see
[Adding tables and figures](adding-content.html)). The running **header** and
**footer** -- the lines repeated at the top and bottom of every page -- are not
attached to individual pages. Instead they are defined by **sections**.

## The model: sections overlay headers/footers onto page ranges

A **section** assigns a header and footer starting at a given page; that
header/footer applies to every page **from there until the next section**. There
is no separate "per-page header" mechanism -- a single page''s header is simply
whatever the section covering that page defines. You create sections with
`rtf_section(doc, page = , secinfo = list(header = , footer = ))`.

Two consequences worth remembering:

- Passing `header = NULL` (or `footer = NULL`) for a section means **inherit**
  the previous section''s header (footer) -- not "blank".
- Sections are an overlay on the flat page list; they do **not** contain or
  group the content. Content order alone determines page numbers.


## Building a header or footer

`rtf_header()` / `rtf_footer()` build the header/footer content. Rows are named
character vectors using `l` / `c` / `r` for the left / centre / right cell; pass
a `list` of them for a multi-row header. Page-number tokens are substituted at
render time:

- `{AUTO_PAGE}` -- the current page number (updates in the viewer);
- `{AUTO_TOTAL_PAGES}` -- the total page count;
- `{PAGE}` -- the static first-page number of the section.

```{r build-hf}
hdr <- rtf_header(rows = list(
  c(l = "Protocol XYZ-001", r = "Confidential"),
  c(l = "Table 14.1.1",     r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}")
))
ftr <- rtf_footer(c(c = "ACME Pharma, Inc."))
```

`update_header_row()` / `update_footer_row()` add or replace one row after
creation (handy when a row depends on logic):

```{r update-row}
hdr <- update_header_row(hdr, row = 3, content = c(c = "DRAFT"))
```

A header/footer can also take a `border` (e.g. a rule under the footer) -- see
`?rtf_header`.


## One header/footer for the whole document

For a single running header/footer, define **one** section starting at page 1:

```{r whole-doc}
df <- data.frame(A = c("1", "2"), B = c("x", "y"), stringsAsFactors = FALSE)

doc <- rtf_document() |>
  rtf_tables(list(df, df, df)) |>                      # three pages
  rtf_section(page = 1, secinfo = list(header = hdr, footer = ftr))
```

The section at page 1 covers all subsequent pages, so every page gets the same
header and footer.


## Different headers/footers for different page ranges

Define more sections, each starting where its header/footer should take over.
Here pages 1-2 use one header and pages 3+ use another:

```{r ranges}
hdr_b <- rtf_header(rows = list(c(l = "Protocol XYZ-001", r = "Table 14.2.1")))

doc <- rtf_document() |>
  rtf_tables(list(df, df, df, df)) |>
  rtf_section(page = 1, secinfo = list(header = hdr,   footer = ftr)) |>
  rtf_section(page = 3, secinfo = list(header = hdr_b, footer = NULL))
#                                                       ^ NULL = keep ftr
```

`footer = NULL` on the second section inherits the first section''s footer, so
only the header changes at page 3. You can also assign several sections in one
call by passing a vector of `page`s and a list of `secinfo` objects -- see
`?rtf_section`.


## Per-content automatic headers -- `auto_section`

When each table should start its own section with a heading derived from its
name, use `auto_section = TRUE` with a **named** list of content. Define the
common header once with `rtf_section()` *without* a `page` (the base template),
and each name is appended to it as a per-section line:

```{r auto-section}
doc <- rtf_document() |>
  rtf_section(secinfo = list(header = hdr, footer = ftr)) |>   # base (no page)
  rtf_tables(
    list("Demographics" = df, "Adverse events" = df),
    auto_section        = TRUE,
    section_label_align = "center"
  )
```

This splits the document into one section per named item, each carrying the base
header plus its own label -- a compact way to get per-table headings without
writing a `rtf_section()` call for every table. `auto_section` is sugar built on
the same section mechanism described above.


## Page numbering

The tokens render to dynamic RTF fields, so the page numbers are correct in the
word processor even after the document is re-paginated:

```{r numbering}
doc <- rtf_document() |>
  rtf_tables(list(df, df)) |>
  rtf_section(page = 1, secinfo = list(
    header = NULL,
    footer = rtf_footer(c(c = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"))
  ))

rtf <- {
  f <- tempfile(fileext = ".rtf"); generate_rtfreport(doc, f, overwrite = TRUE)
  paste(readLines(f, warn = FALSE), collapse = "\n")
}
grepl("\\chpgn", rtf, fixed = TRUE)        # {AUTO_PAGE} -> \chpgn field
```

See `?rtf_section`, `?rtf_header`, and `?rtf_footer` for the full reference.
