---
title: "Importing tables with as_rtftables()"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
.has_gt <- requireNamespace("gt", quietly = TRUE)
.has_flextable <- requireNamespace("flextable", quietly = TRUE)
```

```{r setup}
library(rtfreporter)
```

`as_rtftables()` is the single entry point for turning a **table object** into
the `rtftable` page objects that `rtf_tables()` consumes. It does two jobs:

1. **Read** the source table -- its rendered body plus the metadata the RTF
   renderer can use (column labels, alignment, spanning headers, titles,
   footnotes, …);
2. **Paginate** -- split the body into one `rtftable` per page.

This article covers the **reading / conversion** side. Pagination has its own
guide: see [Paginating with as_rtftables()](pagination.html).

```{r intro}
df <- data.frame(
  Parameter = c("Age (years)", "  Mean", "  SD"),
  Placebo   = c("", "75.1", "8.2"),
  Active    = c("", "74.4", "7.9"),
  stringsAsFactors = FALSE
)

pages <- as_rtftables(df)     # a list of rtftable page objects
length(pages)
class(pages[[1]])
```


## Supported inputs

`as_rtftables()` accepts any of:

- a **`gt`** table (`gt::gt(...)`),
- a **gtsummary** table (`tbl_summary()`, `tbl_regression()`, …; converted via
  `gtsummary::as_gt()`),
- an **rtables / tern** `VTableTree`,
- a **`flextable`** table (`flextable::flextable(...)`),
- a plain **`data.frame` / tibble**, or
- a **`list`** of any of these (flattened to one page each, names propagated).

It always returns a **list** of `rtftable` objects. For a single table where you
do not need a list, use the singular convenience wrapper `as_rtftable()` -- it is
exactly `as_rtftables(x, split = "none")[[1]]`:

```{r singular}
tbl <- as_rtftable(df)        # one rtftable, not a list
class(tbl)
```


## Reading metadata -- `read_meta`

For a rich source (gt / gtsummary / rtables / flextable), `as_rtftables()` reads
the render-relevant metadata so you do not have to restate it. What is carried:

| Metadata | gt / gtsummary | rtables / tern | flextable |
|----------|----------------|----------------|-----------|
| Column (leaf) labels | yes | yes | yes |
| Per-column alignment | yes | yes | yes |
| Spanning headers | yes | yes | yes |
| Column widths | yes (px/pct) | -- | -- |
| Title + subtitle | yes | yes | yes (caption) |
| Footnotes / source notes | yes | yes | yes (footer) |
| In-cell footnote marks | yes (superscript) | yes (superscript) | -- |
| Row-group rows + indent | yes (rendered) | yes (rendered) | yes (rendered) |

For **flextable** the *displayed* text is read (header labels from
`set_header_labels()` and `colformat_*()` formatting included), not the raw
`$body$dataset`. Image/equation cells and `footnote()` reference marks are not
carried.

**Not carried** (RTF cannot reproduce them, so they are intentionally ignored):
per-cell bold / italic / underline, cell background fills, font/size styling, and
Markdown inside labels or titles. A plain `data.frame` / tibble carries **no**
metadata -- you set `col_header`, `col_spec`, etc. yourself (see *Pass-through*
below).

`read_meta` controls the extraction:

- `TRUE` (default) -- read everything in the table above.
- `FALSE` -- use only the rendered body (ignore labels / titles / notes).
- a **character vector of tokens** -- selective opt-in, e.g.
  `c("col_header", "alignment")`. Tokens for gt/gtsummary: `"col_header"`,
  `"alignment"`, `"spanning"`, `"widths"`, `"titles"`, `"footnotes"`. For
  rtables/tern: `"col_header"`, `"alignment"`, `"spanning"`, `"titles"`,
  `"footnotes"`, `"indent"`, `"footnote_marks"`. For flextable:
  `"col_header"`, `"alignment"`, `"spanning"`, `"titles"`, `"footnotes"`.

```{r gt-meta, eval = .has_gt}
library(gt)

g <- gt(head(mtcars[, c("mpg", "cyl")], 3)) |>
  cols_label(mpg = "Miles/gallon", cyl = "Cylinders") |>
  cols_align("right", columns = c(mpg, cyl))

tbl <- as_rtftable(g)               # reads the labels + alignment
tbl$col_header                       # -> "Miles/gallon", "Cylinders"
```

```{r gt-meta-off, eval = .has_gt}
# read_meta = FALSE keeps only the rendered body and ignores the labels, so no
# col_header metadata is extracted (the body's own column names are used).
plain <- as_rtftable(g, read_meta = FALSE)
plain$col_header        # NULL -- nothing extracted
names(plain$data)       # the body column names that headers fall back to
```

A **flextable** is read the same way. The *displayed* text is taken (so
`set_header_labels()` relabelling and `colformat_*()` formatting are honoured),
along with the caption (title), footer lines (footnotes), spanning headers and
per-column alignment:

```{r flextable-meta, eval = .has_flextable}
library(flextable)

ft <- flextable(data.frame(Param = c("Mean", "SD"), Value = c(75.13, 8.234))) |>
  set_header_labels(Param = "Characteristic") |>
  colformat_double(j = "Value", digits = 1) |>
  align(j = "Value", align = "right", part = "body") |>
  set_caption("Table 1: Demographics")

tbl <- as_rtftable(ft)
tbl$data[[2]]            # -> "75.1", "8.2"  (colformat applied)
attr(tbl, "rtf_titles") # -> "Table 1: Demographics"  (from the caption)
```


## Monospaced alignment -- `cell_format`

Clinical tables often need count/percent columns to line up by character.
`cell_format` re-formats the body **before** pagination, column by column. Pass a
single function (applied to every data column; the row-label column 1 is left
alone) or a list of functions taken positionally. rtfreporter ships a few:

- `fmt_count_paren()` -- right-justify an integer count and the number inside the
  parentheses, e.g. `"10 (11.6%)"` over `" 4 ( 4.7%)"`;
- `fmt_right_align()` -- the minimal "right-justify a column" formatter and a
  template for writing your own;
- `align_count_pct = TRUE` -- the long-standing shorthand for the built-in
  `"n (xx.x)"` realigner.

```{r cellfmt}
counts <- data.frame(
  Group = c("A", "B"),
  n_pct = c("4 (4.7%)", "10 (11.6%)"),
  stringsAsFactors = FALSE
)
tbl <- as_rtftable(counts, cell_format = fmt_count_paren)
tbl$data$n_pct                       # both cells padded to one width
```

### Writing your own `cell_format` function

When none of the built-ins matches your data's notation, write your own. The
contract is small:

* **Signature.** The framework calls your function with **one argument** -- a
  character vector holding the cells of a single column. (The built-ins also
  accept a second `nbsp =` argument with a default, but `cell_format` never
  passes it, so a plain `function(x)` is enough.) Your function is called **once
  per column**, and is not told *which* column it is, so base every decision on
  `x` alone.
* **Return value.** A character vector of the **same length** as `x` -- one
  element per row. Returning a different length is an error; never drop or add
  rows.
* **Leave non-matching cells alone.** Cells you do not intend to touch -- empty
  group-label cells, totals, anything not matching your pattern -- must be
  returned unchanged.
* **Pad with the non-breaking space `" "`, not a normal space.** RTF and
  Word collapse runs of ordinary spaces, which would silently undo your
  alignment.
* **Which columns it sees.** A *single* function is applied to columns
  `2..ncol` (column 1 is the row label, left alone by clinical convention). To
  target specific columns, pass a **list** instead: `cell_format[[j]]` formats
  column `j`, and `NULL` entries leave a column untouched.

Here is a complete formatter that appends a unit and right-aligns a numeric
column, leaving blanks and non-numbers as they are:

```{r cellfmt-custom}
# Right-align a numeric column and append a unit (e.g. "mg"), padding on the
# left with non-breaking spaces so the values line up in a monospaced font.
fmt_with_unit <- function(x, unit = "mg", nbsp = " ") {
  x <- as.character(x)
  is_num <- grepl("^[0-9]+(\\.[0-9]+)?$", trimws(x))   # which cells to touch
  labelled <- x
  labelled[is_num] <- paste0(trimws(x[is_num]), " ", unit)
  width <- max(nchar(labelled[is_num]), 0L)            # widest labelled cell
  out <- x
  out[is_num] <- formatC(labelled[is_num], width = width)   # right-justify
  gsub(" ", nbsp, out, fixed = TRUE)                   # protect the padding
}

dose <- data.frame(
  Arm  = c("Low", "High", "Total"),
  Dose = c("5", "12.5", ""),                           # blank cell left as-is
  stringsAsFactors = FALSE
)

# A list targets one column by position; column 1 (Arm) is left untouched.
tbl <- as_rtftable(
  dose,
  cell_format = list(NULL, function(x) fmt_with_unit(x, unit = "mg"))
)
tbl$data$Dose
```

Because `cell_format` passes only the column vector, wrap a parameterised
formatter in a one-argument closure (`function(x) fmt_with_unit(x, unit = "mg")`)
to fix its extra arguments -- exactly as shown above.


## Column widths -- `auto_width`

By default columns share the width evenly. With `auto_width = TRUE` each column
is sized to its widest content (header label or data cell) so long labels do not
wrap; the widths are computed once and applied to every page:

```{r autowidth}
tbl <- as_rtftable(df, auto_width = TRUE)
tbl$column_widths_twips
```

Add `table_width_twips =` to scale those widths to a total (e.g. to fill, or fit
within, the writable page width). An explicit `column_widths_twips` /
`col_rel_width` always wins over `auto_width`.


## Pass-through to `rtftable()`

Any further argument is forwarded to [rtftable()] for **every** page, so you can
set the table''s appearance in the same call -- and your explicit values always
win over the metadata extracted from the source:

```{r passthrough}
tbl <- as_rtftable(
  df,
  border        = "tfl",
  col_header    = c("Parameter", "Placebo (N=86)", "Active (N=84)"),
  col_rel_width = c(2, 1, 1),
  row_height_twips = 280L
)
tbl$col_header
```

See the [page & document setup](page-setup.html) article for document-wide
settings, and `?rtftable` for the full list of per-table options (alignment,
per-column/-cell styles and colours, blank rows, …).


## Titles and footnotes

When the source carries a title / subtitle or source notes, `as_rtftables()`
attaches them to each page as the `rtf_titles` / `rtf_footnotes` attributes, and
`rtf_tables()` renders them automatically -- you do not pass them again:

```{r titles, eval = FALSE}
doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(as_rtftables(my_gt_table))   # gt's title/source notes flow through
```

For a plain data.frame (no metadata), set them on `rtf_tables()` with `titles =`
/ `footnotes =` instead.


## Pagination

Everything above is independent of how the table is split into pages. When the
body is longer than one page, control the split with `as_rtftables()`s
`split` / `max_rows` / `group_col` arguments (or a custom function) --
all of that is covered in
[Paginating with as_rtftables()](pagination.html).

See `?as_rtftables` for the complete argument reference and the per-source
*what is carried* table.
