---
title: "Page and document setup"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

Every report starts with `rtf_document()`, which carries the **document-wide**
settings: page geometry, fonts, colours, and the default font size. The
defaults are tuned for clinical TFL output, so you can usually call it with no
arguments -- but each piece is configurable. This article walks through what you
can set at the document level and how it cascades to the tables.

```{r quick}
df <- data.frame(
  Parameter = c("Age (years)", "  Mean", "  SD"),
  Value     = c("", "75.1", "8.2"),
  stringsAsFactors = FALSE
)

# All defaults -- landscape Letter, 9pt Courier
doc <- rtf_document() |>
  rtf_tables(as_rtftables(df))
```


## Two entry points: `rtf_document()` and `rtf_config()`

`rtf_document()` sets everything when you create the document. `rtf_config()`
updates an **existing** document; only the arguments you pass are changed (a
`NULL` argument leaves that setting untouched). They take the same four
parameters: `page`, `font_table`, `color_table`, and `default_format`.

```{r config}
doc <- rtf_document()                                   # defaults
doc <- rtf_config(doc, default_format = list(font_size_half_points = 20L))
doc$document$default_format$font_size_half_points        # -> 20 (10pt)
```


## Page geometry -- `page`

Pass `page = list(...)`. Dimensions and margins are given in **inches**; any key
you omit falls back to its default.

| Key | Meaning | Default |
|-----|---------|---------|
| `orientation` | `"landscape"` or `"portrait"` | `"landscape"` |
| `width_in` | Paper width (inches) | `11` |
| `height_in` | Paper height (inches) | `8.5` |
| `margin_top_in` | Top margin | `0.9` |
| `margin_bottom_in` | Bottom margin | `0.9` |
| `margin_left_in` | Left margin | `0.6` |
| `margin_right_in` | Right margin | `0.6` |

The default is **landscape US Letter** (11 × 8.5 in), the usual clinical TFL
page. For a portrait A4 page with wider margins:

```{r portrait}
doc <- rtf_document(
  page = list(
    orientation    = "portrait",
    width_in       = 8.27, height_in = 11.69,   # A4
    margin_left_in = 0.75, margin_right_in = 0.75,
    margin_top_in  = 1.0,  margin_bottom_in = 1.0
  )
) |>
  rtf_tables(as_rtftables(df))
```

The writable width (paper width minus the left/right margins) is what tables are
sized against, so changing the margins automatically reflows full-width tables.


## Fonts -- `font_table`

`font_table` is a list of fonts; the **first** one is the primary font used
throughout the report. The default is **Courier**, a monospaced font: in clinical
listings, monospacing is what lets count/percent columns line up by character
(see the count/percent formatters in the pagination article). Switch to a
proportional font only if you do not rely on character alignment:

```{r font}
doc <- rtf_document(font_table = list(list(name = "Arial"))) |>
  rtf_tables(as_rtftables(df))
```


## Document font size -- `default_format`

The document font size lives in `default_format` and is expressed in
**half-points** (the unit of the RTF `\fs` command): `18` = 9 pt, `20` = 10 pt,
`24` = 12 pt, and so on. 9 pt is the package default.

```{r fontsize}
doc <- rtf_document(default_format = list(font_size_half_points = 22L)) |>  # 11pt
  rtf_tables(as_rtftables(df))
```

The font size also **drives the default row height**: taller text needs taller
rows, so rtfreporter picks a sensible default row height for the chosen size.

```{r rowheight-default}
# Internal lookup, shown here to illustrate the cascade:
rtfreporter:::.default_row_height_twips(18L)   # 9pt  -> 230 twips
rtfreporter:::.default_row_height_twips(22L)   # 11pt -> 270 twips
```


## Default cell / row height

All table-shaped elements -- page header/footer rows, column-header rows, data
rows, blank separator rows, and footnotes -- share a **font-size-aware default
row height** (in twips; 1 twip = 1/1440 inch = 1/20 point). These defaults live
in `inst/resources/rtfreporter_defaults.R` (admin-tunable, package-wide):

| Font size | Default row height |
|-----------|--------------------|
| 8 pt (`16`) | 210 twips |
| 9 pt (`18`) | 230 twips *(default)* |
| 10 pt (`20`) | 250 twips |
| 11 pt (`22`) | 270 twips |
| 12 pt (`24`) | 290 twips |

You rarely need to touch these. To change heights for a specific table, override
them per table on `rtftable()` / `as_rtftables()`:

```{r rowheight-override}
doc <- rtf_document() |>
  rtf_tables(as_rtftables(
    df,
    row_height_twips        = 280L,   # data rows
    header_row_height_twips = 300L,   # column-header rows
    blank_row_height_twips  = 140L    # blank separator rows (e.g. half height)
  ))
```

Cell padding (the gap between the cell border and its text) defaults to `0`
(text sits flush against the border, the typical clinical look) and is likewise
configurable per table via `cell_padding_left_twips` / `cell_padding_right_twips`.


## Colours -- `color_table`

`color_table` is a vector of hex colours that make up the document's palette;
border colours reference it. The default is a single black entry. Add colours
you intend to use (for example a coloured rule line):

```{r colors}
doc <- rtf_document(color_table = c("#000000", "#1F4E79")) |>  # black + navy
  rtf_tables(as_rtftables(df))
```


## A complete example

Putting it together -- a portrait A4 page, Arial at 11 pt, custom margins, and a
two-colour palette:

```{r full, eval = FALSE}
doc <- rtf_document(
  page = list(
    orientation    = "portrait",
    width_in       = 8.27, height_in = 11.69,
    margin_left_in = 0.75, margin_right_in = 0.75,
    margin_top_in  = 1.0,  margin_bottom_in = 1.0
  ),
  font_table     = list(list(name = "Arial")),
  color_table    = c("#000000", "#1F4E79"),
  default_format = list(font_size_half_points = 22L)
) |>
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(c(l = "Protocol XYZ-001", r = "Confidential"))),
    footer = rtf_footer(c(c = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"))
  )) |>
  rtf_tables(as_rtftables(df, row_height_twips = 280L))

generate_rtfreport(doc, "report.rtf", overwrite = TRUE)
```

See `?rtf_document` and `?rtf_config` for the full argument reference.
