---
title: "Paginating tables with as_rtftables()"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

`as_rtftables()` turns a table into a **list of `rtftable` page objects** --
one per RTF page -- ready to hand to `rtf_tables()`. The `split =` argument
decides *how* the rows are grouped into pages. This article walks through the
ways to control that:

1. the **built-in split strategies**,
2. a **custom split function**,
3. **pre-splitting** the data yourself, and
4. an **advanced** hybrid that composes a built-in strategy.

Throughout we use a small demographics-style table. By the clinical-TFL
convention, a row whose first column starts with a non-space character opens a
*group*, and the indented rows below it are that group's members -- which is how
`as_rtftables()` auto-detects groups when `group_col = NULL`.

```{r data}
df <- data.frame(
  Parameter = c("Age (years)", "  Mean", "  SD", "  Min", "  Max",
                "Sex",         "  F", "  M",
                "Race",        "  White", "  Black", "  Asian", "  Other"),
  Value     = c("",     "75.1", "8.2", "50", "89",
                "",     "179 (58%)", "127 (42%)",
                "",     "273 (89%)", "29 (9%)", "2 (1%)", "2 (1%)"),
  stringsAsFactors = FALSE
)
```

A tiny helper to *show* how a result was paginated -- it prints the row-label
column of each returned page:

```{r helper}
show_pages <- function(pages) {
  for (i in seq_along(pages)) {
    nm <- names(pages)[i]
    tag <- if (!is.null(nm) && nzchar(nm)) sprintf(" (%s)", nm) else ""
    cat(sprintf("--- Page %d%s ---\n", i, tag))
    cat(paste0("  ", pages[[i]]$data[[1L]]), sep = "\n")
    cat("\n")
  }
  invisible(pages)
}
```


## 1. Built-in split strategies

Pass `split =` one of the built-in strategy names.

### `"none"` (default) -- one page

The whole table stays on a single page; no row limit is checked.

```{r none}
pages <- as_rtftables(df)            # split = "none"
length(pages)
```

### `"rows"` -- cut at explicit positions

Give `split_rows` the row numbers at which a new page should start. Use this
when you know exactly where the breaks go.

```{r rows}
pages <- as_rtftables(df, split = "rows", split_rows = c(6, 9))
show_pages(pages)
```

### `"group_force"` -- fill `max_rows`, repeat the group label on a cut

Pack up to `max_rows` rows on each page. When a page boundary falls *inside* a
group, a continuation row is inserted at the top of the next page repeating the
group label with a `" (Cont.)"` suffix, so the reader knows the group carries
over.

```{r group_force}
pages <- as_rtftables(df, split = "group_force", max_rows = 6)
show_pages(pages)
```

### `"group_safe"` -- never break a group across pages

Pack *whole* groups onto a page; a group that would overflow is pushed to the
next page instead of being split. (A single group larger than `max_rows` is
still force-split as a fallback.)

```{r group_safe}
pages <- as_rtftables(df, split = "group_safe", max_rows = 6)
show_pages(pages)
```

### `"by_value"` -- one page per group, named by the group

Emit exactly one page per detected group; the returned list is **named** by the
group label (which `rtf_tables()` can use, e.g. with `auto_section = TRUE`).

```{r by_value}
pages <- as_rtftables(df, split = "by_value")
names(pages)
```

### The same strategies as reusable functions

Each string strategy has an exported **factory** that returns the same splitter
as a function. These are interchangeable with the string forms, but you can
configure them once and reuse them, or pass them around:

```{r factories}
strategy <- page_split_group_safe(max_rows = 6, group_col = NULL)
pages    <- as_rtftables(df, split = strategy)
length(pages)
```

The factories are `page_split_none()`, `page_split_rows()`,
`page_split_group_safe()`, `page_split_group_force()`, and
`page_split_by_value()`.

### Shared knobs

These arguments shape pagination regardless of the strategy:

| Argument | Effect |
|----------|--------|
| `max_rows` | Maximum data rows per page (required for the group strategies). |
| `group_col` | Group column: a name, a 1-based index, or `NULL` to auto-detect groups from leading whitespace in the first column. |
| `min_group_rows` | Widow/orphan control: avoid stranding a lone group header (or too few of its rows) at a page boundary. Set `0` to disable. |
| `cont_label` | The continuation suffix (`" (Cont.)"`) repeated on a mid-group cut. |


## 2. A custom split function

When the built-ins don't fit, pass a **function** as `split =`. It is called on
the (cell-formatted) body and must return a **list of data.frames** -- one per
page. Named list elements become page names, exactly like `"by_value"`.

```{r custom-basic}
# Fixed blocks of four rows, ignoring groups entirely.
by_four <- function(df, ...) {
  split(df, (seq_len(nrow(df)) - 1L) %/% 4L)
}
pages <- as_rtftables(df, split = by_four)
length(pages)
```

Your function only implements the split; the rest of the pipeline (blank rows,
metadata, per-page assembly, and header/width/style replication) is applied to
its output unchanged. Write it with a `...` so it tolerates the context
arguments (`max_rows`, `group_col`, `cont_label`, `min_group_rows`) it does not
use.

To re-create the `" (Cont.)"` continuation row the built-in group strategies
add, use `add_cont_label()`:

```{r custom-cont}
by_four_cont <- function(df, cont_label = " (Cont.)", ...) {
  parts <- split(df, (seq_len(nrow(df)) - 1L) %/% 4L)
  for (i in seq_along(parts)) {
    if (i > 1L) parts[[i]] <- add_cont_label(parts[[i]], "…continued",
                                             cont_label = cont_label)
  }
  parts
}
show_pages(as_rtftables(df, split = by_four_cont))
```


## 3. Pre-splitting the data yourself

Sometimes the simplest thing is to split the data.frame *before* calling
`as_rtftables()` and hand it a list. `as_rtftables()` flattens a list input into
one page per element, carrying the list's names through as page names.

```{r presplit}
top    <- df[1:8, ]
bottom <- df[9:13, ]

pages <- as_rtftables(list("Demographics (1 of 2)" = top,
                           "Demographics (2 of 2)" = bottom))
names(pages)
```

This is handy when the split rule lives in your own data-wrangling code (for
example `split(df, df$SOC)` for a list of adverse-event tables), or when each
page comes from a different source entirely.


## 4. Advanced: composing a built-in strategy

The factories from section 1 are ordinary functions, so you can call one inside
your own splitter and then post-process its output -- a hybrid of approaches 1
and 2. For example, reuse `page_split_group_safe()` but drop any empty pages and
relabel continuations:

```{r compose}
safe_trimmed <- function(df, ...) {
  parts <- page_split_group_safe(max_rows = 6, group_col = NULL)(df)
  Filter(function(p) nrow(p) > 0L, parts)   # keep only non-empty pages
}
length(as_rtftables(df, split = safe_trimmed))
```


## Blank separator rows

Clinical tables often need **blank separator rows** -- empty rows that put a
little air between groups. They are not part of the data; they are layout, added
on top of the body.

### Blank rows and the row count

This matters for pagination: **`max_rows` counts data rows only.** Blank rows are
inserted into each page *after* the split, so they do not use up the `max_rows`
budget. For example, packing whole groups with `max_rows = 6` and a blank
between groups:

```{r blank-count}
df_g <- data.frame(
  label = c("Group A", "  a1", "  a2", "Group B", "  b1", "  b2",
            "Group C", "  c1", "  c2"),
  v = as.character(1:9), stringsAsFactors = FALSE
)

pages <- as_rtftables(df_g, split = "group_safe", max_rows = 6,
                      blank_rows = "between_groups")
nrow(pages[[1]]$data)          # 6 data rows -- the max_rows budget
pages[[1]]$blank_rows          # plus a blank after row 3 (A -> B), not counted
```

So page 1 shows six data rows **and** a separator -- seven rows on the page --
but only the six data rows counted toward `max_rows`.

### Adding blank rows while paginating

`as_rtftables()` takes the blank-row controls alongside the split arguments;
they are applied to every page:

- `blank_rows = "between_groups"` -- a blank at each group transition on a page;
- `blank_rows = c(0, 3)` -- explicit positions (`0` = before the first row,
  `k` = after row `k`), relative to each page;
- `blank_row_first = TRUE` / `blank_row_end = TRUE` -- a blank at the top /
  bottom of every page.

### Paginating yourself: `set_blank_rows()`

`set_blank_rows()` is the helper `as_rtftables()` calls on every chunk. If you do
your own splitting, use it to attach the blank positions to a page-sized
data.frame; `rtftable(read_attributes = TRUE)` (the default) then picks them up:

```{r set-blank}
page <- set_blank_rows(df_g, blank_rows = "between_groups")
attr(page, "rtf_blank_rows")
tbl <- rtftable(page)          # read_attributes = TRUE consumes the attribute
```

### Blank rows on a single (un-paginated) table

You do not need pagination to use them: `rtftable(blank_rows = )` accepts integer
positions (here `-1` also means "after the last row"), or the spec constructors
[blank_rows_by_change()] (blank when a column value changes) and
[blank_rows_by_rule()] (blank before/after rows matching a regex), or a `list`
combining them:

```{r blank-single}
rtftable(df_g, blank_rows = blank_rows_by_change("label"))
```

Control the separator height with `blank_row_height_twips` (e.g. half the data
row height); see [Page & document setup](page-setup.html). The full spec forms
are documented in `?rtftable` and `?set_blank_rows`.


## Choosing an approach

| You want… | Use |
|-----------|-----|
| Everything on one page | `split = "none"` (default) |
| Breaks at known row numbers | `split = "rows"` + `split_rows` |
| As many rows as fit, groups may split (with a continuation label) | `split = "group_force"` + `max_rows` |
| As many *whole groups* as fit, never split a group | `split = "group_safe"` + `max_rows` |
| One page per group, named | `split = "by_value"` |
| A bespoke rule | a custom `split =` function (+ `add_cont_label()`) |
| The split already done upstream | pass a (named) list to `as_rtftables()` |
| A built-in rule plus a tweak | wrap a `page_split_*()` factory |

Whichever you choose, the result is the same shape -- a list of `rtftable`
pages -- so the rest of your pipeline does not change:

```{r pipeline, eval = FALSE}
doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(as_rtftables(df, split = "group_safe", max_rows = 6))

generate_rtfreport(doc, "demographics.rtf", overwrite = TRUE)
```

See `?as_rtftables` for the full `split =` contract and `?page_split` for the
factory functions.
