---
title: "Same report, every table framework"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE, eval = FALSE)

# Show a screenshot of the rendered RTF if it has been captured.  The PNGs sit
# next to the .rtf files in inst/rtf-examples/showcase/; we copy the one we need
# into the article's own figures/ folder (so pkgdown serves it) and include it.
# If the screenshot does not exist yet, a small placeholder is shown instead so
# the article still builds.
.snapshot <- function(file, alt = "rendered RTF") {
  src <- file.path("..", "..", "inst", "rtf-examples", "showcase", file)
  if (file.exists(src)) {
    dir.create(file.path("figures", "showcase"), showWarnings = FALSE,
               recursive = TRUE)
    dest <- file.path("figures", "showcase", file)
    file.copy(src, dest, overwrite = TRUE)
    knitr::include_graphics(dest)
  } else {
    knitr::asis_output(sprintf(
      "<p><em>(snapshot <code>inst/rtf-examples/showcase/%s</code> not captured yet)</em></p>",
      file))
  }
}
```

```{r setup}
library(rtfreporter)
```

The R ecosystem for clinical tables is wonderfully diverse -- **rtables/tern**,
**tfrmt**, **gtsummary** (on its own or on a **cards/cardx** ARD), **Tplyr**,
**gt**, **flextable**, **huxtable** -- and teams pick whichever fits their style
and validation story. What they all have in common is the *last* step: a
**regulatory RTF** that opens cleanly in Word.

That last step is exactly what `rtfreporter` does, for **all of them**. This
article builds the **same demographics report from the same data** with several
frameworks and renders each to RTF through `rtfreporter`. The point is not that
one framework is best -- it is that **whatever you already use, rtfreporter is
the RTF back end**.

> **Every block here is copy-paste runnable, from data creation to the written
> `.rtf`.** Run a section's **data** and **furniture** blocks once (they prepare
> the data and the shared rtfreporter header / footer / column spec), then copy
> any single framework block: it builds that framework's table and then calls
> rtfreporter **directly** -- `as_rtftables()`, `rtf_document()`,
> `generate_rtfreport()` -- to write its `.rtf`. We render with `eval = FALSE` in the
> article itself -- so pkgdown does not have to run the entire framework stack --
> and show the Word output as a screenshot. The same generators also live in
> [`data-raw/showcase_dm.R`](https://github.com/ichirio/rtfreporter/blob/main/data-raw/showcase_dm.R)
> and
> [`data-raw/showcase_ae.R`](https://github.com/ichirio/rtfreporter/blob/main/data-raw/showcase_ae.R),
> and the rendered `.rtf` files are committed to
> [`inst/rtf-examples/showcase/`](https://github.com/ichirio/rtfreporter/tree/main/inst/rtf-examples/showcase).

# Demographics (DM)

## The data

We use [`pharmaverseadam`](https://pharmaverse.github.io/pharmaverseadam/)'s
`adsl`, restricted to the three randomised arms, and summarise **age** (as `n`,
`Mean (SD)`, `Median`, `Min, Max`), an **age category**, **sex** and **race** by
treatment -- a production-style Table 14.1 demographics layout.

```{r data}
library(dplyr)
arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl |>
  filter(TRT01A %in% arm_levels) |>
  mutate(
    TRT01A = factor(TRT01A, levels = arm_levels),
    SEX    = factor(SEX, levels = c("M", "F"), labels = c("Male", "Female")),
    AGEGR  = cut(AGE, breaks = c(-Inf, 65, 81, Inf),
                 labels = c("<65", "65 - 80", ">80"), right = FALSE),
    # All race categories as a factor, so even zero-count levels (Asian,
    # American Indian) appear as "0 (0.0%)" in every framework.
    RACE   = factor(RACE, levels = c(
      "WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN",
      "AMERICAN INDIAN OR ALASKA NATIVE"),
      labels = c("White", "Black or African American", "Asian",
                 "American Indian or Alaska Native")))
arm_n <- table(adsl$TRT01A)
```

## The shared rtfreporter furniture, defined once

The running **header** (sponsor, protocol, page number and the table title), the
**footer** (analysis note, program path and run time), the **Letter landscape**
page, the identical column header with the arm Ns, the 40 : 20 : 20 : 20 column
widths and the left / centred alignment are all rtfreporter settings -- the same
for every framework below. We build them **once here**, as plain rtfreporter
objects and values, and every framework block then references them. Only the
table *body* changes from framework to framework.

```{r furniture}
# Running header: sponsor / protocol / page number, then the title block.
dm_header <- rtf_header(rows = list(
  c(l = "Acme Biopharma, Inc.", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(l = "Protocol ABC-2026-001", r = "Status: Draft"),
  c(c = "Table 14.1.1"),
  c(c = "Summary of Demographic and Baseline Characteristics"),
  c(c = "Safety Population"),
  c(c = "")))                                       # gap before the content

# Footer: analysis note, program path and run time.
dm_footer <- rtf_footer(rows = list(
  c(l = "Note: Percentages are based on the number of non-missing subjects in each treatment group."),
  c(l = "Program: /prod/abc/tfl/t_14_1_1_dm.R", r = "Generated: 2026-06-17 09:14"),
  c(l = "Source: ADSL", r = "CONFIDENTIAL")))

# One identical column header for every framework (arm Ns underneath), the
# 40 : 20 : 20 : 20 relative widths, and row labels left / data columns centred.
dm_col_header <- c("Characteristic",
                   paste0(arm_levels, "\nN = ", as.integer(arm_n)))
dm_col_spec <- c(
  list(list(col = 1L, align = "left",   header_align = "center")),
  lapply(2:4, function(j) list(col = j, align = "center", header_align = "center")))
dm_widths <- c(40, 20, 20, 20)
```

Every framework block below ends with the **same three rtfreporter steps**,
written out in full so nothing is hidden:

1. `as_rtftables()` turns the framework's table into a list of rtfreporter pages,
   taking the shared `dm_col_header` / `dm_col_spec` / `dm_widths` and a blank row
   between groups (`blank_rows = "between_groups"`);
2. `rtf_document() |> rtf_section() |> rtf_tables()` wraps those pages with the
   shared `dm_header` and `dm_footer` on a Letter-landscape page;
3. `generate_rtfreport()` writes the `.rtf`.

## A table per framework

Each framework below builds the demographics **body** its own way; we then hand
that object straight to the shared rtfreporter pipeline. `as_rtftables()` reads
the rendered body (and, with `read_meta = TRUE`, any structural metadata), while
our explicit `dm_col_header` / `dm_col_spec` / `dm_widths` force the one identical
regulatory look across all of them.

### gtsummary

```{r gtsummary}
library(gtsummary)
tbl <- adsl |>
  select(TRT01A, AGE, AGEGR, SEX, RACE) |>
  tbl_summary(
    by = TRT01A,
    type = list(AGE ~ "continuous2"),                       # the 4-line age block
    label = list(AGE = "Age (years)", AGEGR = "Age category",
                 SEX = "Sex", RACE = "Race"),
    statistic = list(
      AGE ~ c("{N_nonmiss}", "{mean} ({sd})", "{median}", "{min}, {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(AGE ~ list(N_nonmiss = 0, mean = 1, sd = 2,
                             median = 1, min = 0, max = 0)))
# (No modify_header: rtfreporter supplies the shared column header below.)

# align_count_pct lines up the "n (xx.x%)" cells; the shared furniture supplies
# the header / spec / widths.  Now render straight to RTF:
pages <- as_rtftables(tbl, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_gtsummary.rtf", overwrite = TRUE)
```

```{r img-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_gtsummary.png", "DM summary via gtsummary")
```

### gtsummary on a cards / cardx ARD

The modern, ARD-first workflow: compute an **Analysis Results Dataset** with
`cards`, summarise it with `gtsummary::tbl_ard_summary()`, then hand the table
to `rtfreporter` exactly as above.

```{r gtsummary-ard}
library(cards); library(gtsummary)
ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE,
                 statistic = ~ list(N = length, mean = mean, sd = sd,
                                    median = median, min = min, max = max)),
  ard_categorical(variables = c(AGEGR, SEX, RACE)), .total_n = TRUE)

tbl <- tbl_ard_summary(
  ard, by = TRT01A,
  type = list(AGE = "continuous2"),
  label = list(AGE = "Age (years)", AGEGR = "Age category",
               SEX = "Sex", RACE = "Race"),
  statistic = list(AGE = c("{N}", "{mean} ({sd})", "{median}", "{min}, {max}"),
                   all_categorical() ~ "{n} ({p}%)"))

pages <- as_rtftables(tbl, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_gtsummary_ard.rtf", overwrite = TRUE)
```

```{r img-gtsummary-ard, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_gtsummary_ard.png", "DM summary via gtsummary on a cards/cardx ARD")
```

### rtables / tern

`tern::analyze_vars()` produces a `VTableTree`; `rtfreporter` reads its row
labels, column counts and spanning structure via `formatters::matrix_form()`.

```{r rtables}
library(rtables); library(tern)
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze_vars(
    vars = c("AGE", "AGEGR", "SEX", "RACE"),
    var_labels = c(AGE = "Age (years)", AGEGR = "Age category",
                   SEX = "Sex", RACE = "Race"),
    .stats = c("n", "mean_sd", "median", "range", "count_fraction"),
    # render the count/percent cells as "n (xx.x%)" to match the other frameworks.
    .formats = c(count_fraction = function(x) {
      sprintf("%d (%.1f%%)", round(x[1]), 100 * x[2])
    }))
tbl <- build_table(lyt, adsl)

pages <- as_rtftables(tbl, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_rtables.rtf", overwrite = TRUE)
```

```{r img-rtables, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_rtables.png", "DM summary via rtables/tern")
```

### cards + tfrmt

The *same* cards ARD as above, rendered with **tfrmt** instead of gtsummary --
so the one ARD drives two different formatters. The ARD is reshaped to tfrmt's
`group / label / column / param / value` long form, a `tfrmt` spec formats it,
and `print_to_gt()` produces the `gt` that `rtfreporter` reads.

```{r cards-tfrmt}
library(cards); library(tfrmt)
ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE,
                 statistic = ~ list(n = length, Mean = mean, SD = sd,
                                    Median = median, Min = min, Max = max)),
  ard_categorical(variables = c(AGEGR, SEX, RACE)))

# Reshape the ARD to tfrmt's long form: group / label / column / param / value.
# sc1()/nm1() pull the first element of each list-column as character / numeric.
d   <- as.data.frame(ard)
sc1 <- function(x) vapply(x, function(v) if (length(v)) as.character(v[[1]]) else NA_character_, character(1))
nm1 <- function(x) vapply(x, function(v) if (length(v)) suppressWarnings(as.numeric(v[[1]])) else NA_real_, numeric(1))

long <- d |>
  mutate(column = sc1(group1_level), vlev = sc1(variable_level), value = nm1(stat)) |>
  filter((variable == "AGE" & stat_name %in% c("n", "Mean", "SD", "Median", "Min", "Max")) |
         (variable %in% c("AGEGR", "SEX", "RACE") & stat_name %in% c("n", "p"))) |>
  transmute(
    column = factor(column, levels = arm_levels),
    group  = recode(variable, AGE = "Age (years)", AGEGR = "Age category",
                    SEX = "Sex", RACE = "Race"),
    label  = case_when(
      variable == "AGE" & stat_name == "n" ~ "n",
      variable == "AGE" & stat_name %in% c("Mean", "SD") ~ "Mean (SD)",
      variable == "AGE" & stat_name == "Median" ~ "Median",
      variable == "AGE" & stat_name %in% c("Min", "Max") ~ "Min, Max",
      TRUE ~ vlev),
    param  = ifelse(stat_name == "p", "pct", stat_name),
    value  = ifelse(stat_name == "p", value * 100, value))

# Everything except the categorical "n (pct%)" cell is shared; only the
# count/percent formatting -- and who aligns it -- changes between the two
# renders below.
tfrmt_spec <- function(count_pct) tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  body_plan = body_plan(
    frmt_structure(".default", ".default", frmt("xx.x")),
    frmt_structure("Age (years)", "n", frmt("xx")),
    frmt_structure(".default", "Mean (SD)",
                   frmt_combine("{Mean} ({SD})",
                                Mean = frmt("xx.x"), SD = frmt("xx.xx"))),
    frmt_structure(".default", "Min, Max",
                   frmt_combine("{Min}, {Max}",
                                Min = frmt("xx"), Max = frmt("xx"))),
    frmt_structure(c("Age category", "Sex", "Race"), ".default", count_pct)))
```

#### Who formats the `n (xx.x%)` cell?

This is a nice place to see the same table formatted **two ways**, because the
count/percent cell is exactly where formatters differ. (Both render chunks
reuse `long` and `tfrmt_spec` from the block just above.)

A small but important detail first: write the percent as **`x.x`**, not `xx.x`.
`frmt("xx.x")` right-pads the integer part, so a one-digit percent prints with a
leading space *inside* the parentheses -- `( 3.5%)`. `frmt("x.x")` gives a clean
`(3.5%)`.

**Way 1 -- tfrmt formats and aligns.** The percent uses `x.x`; the count is
padded to `xx`. `rtfreporter` reads the strings and renders them *as-is*
(`align_count_pct` left off):

```{r tfrmt-self}
g1 <- print_to_gt(
  tfrmt_spec(frmt_combine("{n} ({pct}%)", n = frmt("xx"), pct = frmt("x.x"))),
  long)
# tfrmt already aligned the strings; rtfreporter renders them as-is (no
# align_count_pct).  Direct render:
pages <- as_rtftables(g1, read_meta = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_tfrmt.rtf", overwrite = TRUE)   # tfrmt's alignment
```

```{r img-tfrmt, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_tfrmt.png", "DM summary via cards + tfrmt, formatted by tfrmt")
```

**Way 2 -- rtfreporter formats.** tfrmt does the bare minimum -- just clean
`n (x.x%)` strings, no count padding -- and `rtfreporter` re-aligns them with
`align_count_pct = TRUE`:

```{r tfrmt-rtfreporter}
g2 <- print_to_gt(
  tfrmt_spec(frmt_combine("{n} ({pct}%)", n = frmt("x"), pct = frmt("x.x"))),
  long)
# rtfreporter owns the count/percent alignment via align_count_pct = TRUE.
pages <- as_rtftables(g2, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_tfrmt_rtfreporter.rtf", overwrite = TRUE)  # rtfreporter aligns
```

```{r img-tfrmt-rtfreporter, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_tfrmt_rtfreporter.png", "DM summary via cards + tfrmt, formatted by rtfreporter")
```

The two are deliberately *not* byte-identical. tfrmt anchors the `(` right after
the count, so the percentages stay left-aligned within the parentheses.
`rtfreporter`'s `align_count_pct` instead right-aligns the whole parenthetical so
the **decimal points line up**, and drops the redundant decimal at 100%
(`100%`, not `100.0%`). Use whichever convention your SAP calls for -- the point
is that you can hand `rtfreporter` already-formatted strings *or* let it own the
count/percent alignment, from the very same tfrmt table.

## When the framework returns only the numbers

Some frameworks return just the **formatted body** -- the numbers as strings --
with no layout. You assemble the labelled body yourself, then hand it to the
*same* rtfreporter pipeline as everything else.

### Tplyr

`Tplyr::build()` returns a data frame of formatted strings (plus ordering
helpers). We reshape it into a labelled body -- group headers with indented
stats and a "Number of Subjects" row on top -- and hand that plain data frame to
the shared rtfreporter render.

```{r tplyr}
library(Tplyr)
built <- tplyr_table(adsl, TRT01A) |>
  add_layer(group_desc(AGE, by = "Age (years)") |>
              set_format_strings("n"         = f_str("xx", n),
                                 "Mean (SD)" = f_str("xx.x (xx.xx)", mean, sd),
                                 "Median"    = f_str("xx.x", median),
                                 "Min, Max"  = f_str("xx, xx", min, max))) |>
  # "x.x" (not "xx.x") so a one-digit percent has no leading space inside the
  # parens -- a clean " 1 (1.2%)" that align_count_pct can re-align.
  add_layer(group_count(AGEGR, by = "Age category") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  add_layer(group_count(SEX, by = "Sex") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  add_layer(group_count(RACE, by = "Race") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  build()

# Reshape `built` into `disp`: one column per arm, a label column with each
# group header followed by its indented stat rows, in a fixed display order.
val_cols <- paste0("var1_", arm_levels)
ord <- c("Age (years)", "n", "Mean (SD)", "Median", "Min, Max",
         "Age category", "<65", "65 - 80", ">80",
         "Sex", "Male", "Female",
         "Race", "White", "Black or African American", "Asian",
         "American Indian or Alaska Native")
groups <- c("Age (years)", "Age category", "Sex", "Race")
body <- built[order(built$ord_layer_index, built$ord_layer_1, built$ord_layer_2), ]
disp <- do.call(rbind, lapply(split(body, factor(body$row_label1, groups)), function(g) {
  hdr <- data.frame(Characteristic = g$row_label1[1], stringsAsFactors = FALSE)
  hdr[val_cols] <- ""
  rows <- data.frame(Characteristic = paste0("  ", g$row_label2),
                     stringsAsFactors = FALSE)
  rows[val_cols] <- lapply(val_cols, function(c) g[[c]])
  rbind(hdr, rows)
}))
disp <- disp[order(match(sub("^  ", "", disp$Characteristic), ord)), ]
# "Number of Subjects" line on top (just N).
top <- data.frame(Characteristic = "Number of Subjects", stringsAsFactors = FALSE)
top[val_cols] <- as.character(as.integer(arm_n))
disp <- rbind(top, disp)
names(disp) <- c("Characteristic", arm_levels)
rownames(disp) <- NULL

# Metadata lives HERE, in rtfreporter -- not in Tplyr.  We pass the body plus the
# shared furniture straight to as_rtftables() and render it.
pages <- as_rtftables(disp, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_tplyr.rtf", overwrite = TRUE)
```

```{r img-tplyr, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_tplyr.png", "DM summary via Tplyr")
```

## Two ways to divide the work

The frameworks above split into two camps -- not in how `rtfreporter` renders
them (that pipeline is identical), but in **how much layout the framework hands
you**:

| | **Ready-made body** | **Body you assemble** |
|---|---|---|
| Examples | rtables/tern, gtsummary, cards + gtsummary, cards + tfrmt | Tplyr (and other body-only tools) |
| What the framework gives you | group headers + indented rows, laid out | formatted numbers only |
| Your job before rtfreporter | none -- pass the object | reshape into a labelled body |
| The rtfreporter render | identical | identical |

Either way the regulatory furniture (`dm_col_header` / `dm_col_spec` /
`dm_widths`, plus the shared header and footer) lives **once in rtfreporter**,
and both routes produce the **same RTF**.

# Adverse Events (AE)

The demographics table fits on a page; a real **adverse-events** table does not.
This is where `rtfreporter`'s **pagination** earns its keep: a long
System-Organ-Class / Preferred-Term table is split across pages, and a class
that straddles a page boundary has its label **repeated with `(Cont.)`** at the
top of the next page.

We count the **number of subjects** with at least one treatment-emergent AE,
by treatment group, shown as `n (xx.x%)`:

- **"Subjects with any adverse event"** overall on top, then per **System Organ
  Class (SOC)**, and -- indented -- per **Preferred Term (PT)**.
- Each level is an **independent distinct-subject count**: the SOC count sits on
  the **SOC row itself** and counts a subject once even if they had several PTs
  in that class, so a SOC count is **not** the sum of its PTs.
- Limited to PTs occurring in **>= 3% of subjects in any group** (the overall /
  SOC counts are still over *all* TEAEs), so the table is a readable couple of
  pages.
- **SOCs are listed alphabetically**; within each SOC the **PTs are ordered by
  subject count (all arms), descending**, ties broken **alphabetically**.

## The data, furniture and shared helpers

Run this block once before any AE framework block. It prepares the AE analysis
data (`adae`), the canonical orderings every framework reuses so the tables
match cell-for-cell, the shared column header / spec / widths, the running
header and footer, and a `render_ae()` helper that wraps a paginated list of
pages into the AE document and writes `<name>.rtf`. (It reuses `arm_levels` /
`arm_n` from the DM data block.)

```{r ae-setup}
library(dplyr)
# Treatment-emergent AEs only, in the same three arms as the DM table.
adae <- pharmaverseadam::adae |>
  filter(TRT01A %in% arm_levels, TRTEMFL == "Y") |>
  mutate(TRT01A = factor(TRT01A, levels = arm_levels))
ANY_AE <- "Subjects with any adverse event"

# -- Canonical order -------------------------------------------------------
# SOCs alphabetical; within each SOC, PTs by total distinct subjects (all arms)
# descending, ties A -> Z.  The gtsummary / cards + gtsummary tables get this in
# one native call -- sort_hierarchical() takes a PER-VARIABLE `sort` (one rule
# per hierarchy level); "descending" sorts by the total distinct-subject count
# across all by-groups and breaks ties A -> Z, exactly the order we want.
AE_SORT <- list(AESOC ~ "alphanumeric", AEDECOD ~ "descending")
# The transpose-and-set frameworks (Tplyr, tfrmt) build the body themselves, so
# they need the same PT order as an explicit vector.
.pt_order <- adae |> distinct(USUBJID, AEDECOD) |>
  count(AEDECOD, name = "tot") |>
  arrange(desc(tot), AEDECOD) |> pull(AEDECOD) |> as.character()
# PTs kept in the table: >= 3% of subjects in any treatment group.
.keep_pt <- adae |> distinct(USUBJID, TRT01A, AEDECOD) |>
  count(TRT01A, AEDECOD, name = "n") |>
  mutate(p = 100 * n / as.integer(arm_n[as.character(TRT01A)])) |>
  group_by(AEDECOD) |> summarise(mx = max(p), .groups = "drop") |>
  filter(mx >= 3) |> pull(AEDECOD) |> as.character()
# Real SOC / PT pairs (a PT belongs to exactly one SOC).
.soc_pt <- adae |> distinct(AESOC, AEDECOD) |>
  mutate(across(everything(), as.character))

# -- Shared column header / spec / widths / page budget ---------------------
ae_col_header <- c("System Organ Class /\nPreferred Term",
                   paste0(arm_levels, "\nN = ", as.integer(arm_n)))
ae_col_spec <- c(
  list(list(col = 1L, align = "left",   header_align = "left")),
  lapply(2:4, function(j) list(col = j, align = "center", header_align = "center")))
ae_widths   <- c(0.52, 0.16, 0.16, 0.16)
AE_MAX_ROWS <- 32L

# -- Running header (footer is built per-framework inside render_ae) --------
ae_header <- rtf_header(rows = list(
  c(l = "Acme Biopharma, Inc.",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(l = "Protocol ABC-2026-001", r = "Status: Draft"),
  c(c = "Table 14.3.1"),
  c(c = "Treatment-Emergent Adverse Events Occurring in >= 3% of Subjects in Any Treatment Group"),
  c(c = "by System Organ Class and Preferred Term -- Safety Population"),
  c(c = "")))

# Cell format shared by the read-meta frameworks: collapse a zero cell
# "0 (0.0%)" to a bare "0" (so SOC and PT zeros match tern), then align.
fmt_ae <- function(x, nbsp = " ") {
  z <- grepl("^[[:space:]]*0[[:space:]]*\\([[:space:]]*0(\\.0)?%?[[:space:]]*\\)[[:space:]]*$", x)
  x[z] <- "0"
  fmt_count_paren_bare(x, nbsp = nbsp)
}

# Prepend an "any event" overall row to a tbl_ard_hierarchical() (which, unlike
# tbl_hierarchical(), has no overall_row argument).  This is generic and
# ARD-driven: the overall n / N / p are read straight off an ARD built with
# `cards::ard_stack_hierarchical(..., over_variables = TRUE)` -- the rows whose
# variable is "..ard_hierarchical_overall..".  The by-level -> stat_* column
# mapping comes from the table's own header (`modify_stat_level`), and each cell
# is formatted with the same glue `statistic` template the table uses.  Nothing
# here touches the original data frame.
add_overall_row <- function(tbl, ard, label = ANY_AE,
                            statistic = "{n} ({p}%)", digits = 1L) {
  hdr <- tbl$table_styling$header
  hdr <- hdr[startsWith(hdr$column, "stat_"), ]
  hdr <- hdr[order(hdr$column), ]                  # stat_1, stat_2, ... order
  ov <- ard |>
    dplyr::filter(variable == "..ard_hierarchical_overall..",
                  stat_name %in% c("n", "N", "p")) |>
    dplyr::transmute(level = as.character(unlist(group1_level)),
                     stat_name, stat = unlist(stat)) |>
    tidyr::pivot_wider(names_from = stat_name, values_from = stat)
  cells <- vapply(hdr$modify_stat_level, function(lvl) {
    r <- ov[match(lvl, ov$level), ]
    glue::glue_data(list(n = r$n, N = r$N,
                         p = formatC(100 * r$p, format = "f", digits = digits)),
                    statistic)
  }, character(1))
  tb  <- tbl$table_body
  row <- tb[1, ]; row[seq_len(ncol(row))] <- NA
  row$variable <- "..ard_hierarchical_overall.."
  row$row_type <- "level"
  row$label    <- label
  for (j in seq_len(nrow(hdr))) row[[hdr$column[j]]] <- cells[j]
  tbl$table_body <- dplyr::bind_rows(row, tb)
  # Un-indent the injected row via the public API so it lines up with the SOCs.
  gtsummary::modify_indent(tbl, columns = "label",
                           rows = variable == "..ard_hierarchical_overall..",
                           indent = 0L)
}

# Bake the PT indent into the label TEXT (NBSP), because as_flex_table() /
# as_hux_table() turn gtsummary's indent into cell padding that the flextable /
# huxtable adapters do not read as a row-label indent.
bake_indent <- function(tbl, n = 4L) {
  tb <- tbl$table_body
  pt <- tb$variable == "AEDECOD"
  tb$label[pt] <- paste0(strrep(" ", n), tb$label[pt])
  tbl$table_body <- tb
  tbl$table_styling$indent <- tbl$table_styling$indent[0, ]
  tbl
}

# Wrap a paginated list of rtftable pages into the AE document and write it.
render_ae <- function(pages, name) {
  footer <- rtf_footer(rows = list(
    c(l = "Note: A subject with multiple events within a category is counted once for that category. Percentages use the number of subjects in the safety population per group."),
    c(l = paste0("Program: /prod/abc/tfl/t_14_3_1_", name, ".R"),
      r = "Generated: 2026-06-17 09:14"),
    c(l = "Source: ADSL, ADAE", r = "CONFIDENTIAL")))
  doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
    rtf_section(page = 1, secinfo = list(header = ae_header, footer = footer)) |>
    rtf_tables(pages)
  generate_rtfreport(doc, paste0(name, ".rtf"), overwrite = TRUE)
}
```

## rtables / tern

`tern` is the pharmaverse standard for this table. `analyze_num_patients()`
gives the overall "any AE" row; a SOC-level content summary
(`summarize_row_groups()` with `child_labels = "hidden"`) puts each SOC's
distinct-subject count **on the SOC name row**; `count_occurrences()` gives the
indented PT rows. We build on the full data, **prune** PT rows below 3% with
`rtables::prune_table()`, and **sort** by subject count -- alphabetical factor
levels plus a stable `sort_at_path()` break ties A -> Z. `rtfreporter` then
paginates with **`split = "group_force"`** (`max_rows` per page), which is what
inserts the `(Cont.)` rows.

```{r ae-tern}
library(rtables); library(tern)
# Alphabetical factor levels so SOCs / PTs tie-break A -> Z.
adae_f <- adae |>
  mutate(AESOC   = factor(AESOC,   levels = sort(unique(AESOC))),
         AEDECOD = factor(AEDECOD, levels = sort(unique(AEDECOD))))

# Distinct subjects with any AE in a SOC, per column -- shown on the SOC row.
# A zero count prints as a bare "0" (no parens), exactly like count_occurrences()
# renders the PT rows, so both levels format a zero the same way.
soc_count <- function(df, labelstr, .N_col, ...) {
  n <- length(unique(df$USUBJID))
  cell <- if (n == 0L) rcell(0L, format = "xx")
          else         rcell(c(n, n / .N_col), format = "xx (xx.x%)")
  in_rows(cell, .labels = labelstr)
}

ae <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze_num_patients(vars = "USUBJID", .stats = "unique",
                       .labels = c(unique = ANY_AE)) |>
  split_rows_by("AESOC", child_labels = "hidden", nested = FALSE,
                split_fun = drop_split_levels) |>
  summarize_row_groups(cfun = soc_count) |>              # SOC count on the SOC row
  count_occurrences(vars = "AEDECOD", .indent_mods = 1L) |>
  build_table(df = adae_f, alt_counts_df = adsl) |>
  prune_table(keep_rows(has_fraction_in_any_col(atleast = 0.03))) |>    # >= 3%
  # SOCs stay alphabetical (factor order); only the PTs within each SOC are
  # ordered by subject count (all arms), ties A -> Z.
  sort_at_path(path = c("AESOC", "*", "AEDECOD"), scorefun = score_occurrences)

# group_force = fill each page to max_rows and repeat a split SOC with "(Cont.)"
render_ae(
  as_rtftables(ae, split = "group_force", max_rows = AE_MAX_ROWS,
               blank_rows = "between_groups", cell_format = fmt_count_paren_bare,
               col_header = ae_col_header, col_spec = ae_col_spec,
               col_rel_width = ae_widths, row_height_twips = 200),
  "ae_tern")
```

```{r img-ae-tern, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_tern.png", "AE summary via rtables/tern, paginated with (Cont.)")
```

The same `as_rtftables()` call returns a **list of pages**; `rtf_tables()`
renders them in order, and the running header's `{AUTO_PAGE}` /
`{AUTO_TOTAL_PAGES}` number them.

## gtsummary

`gtsummary::tbl_hierarchical()` builds the **same** table directly:
`overall_row = TRUE` is the any-AE row, each SOC row carries its own subject
count, and PTs are nested beneath. `filter_hierarchical()` keeps the >= 3% PTs,
and `sort_hierarchical()` puts them in the canonical order natively -- its `sort`
takes one rule **per hierarchy level** (`AE_SORT` from setup: SOCs alphabetical,
PTs by subject count, ties A -> Z). The overall row is labelled to match tern via
the native `label` argument, and `fmt_ae()` collapses a zero cell to a bare `0`.

```{r ae-gtsummary}
library(gtsummary)
tbl <- tbl_hierarchical(
    data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
    denominator = adsl, id = USUBJID, overall_row = TRUE,
    label = list("..ard_hierarchical_overall.." ~ ANY_AE),
    statistic = ~ "{n} ({p}%)", digits = everything() ~ list(p = 1)) |>
  filter_hierarchical(p >= 0.03) |>
  sort_hierarchical(sort = AE_SORT)    # SOCs A->Z, PTs by subject count

render_ae(
  as_rtftables(tbl, read_meta = TRUE, split = "group_force", max_rows = AE_MAX_ROWS,
               blank_rows = "between_groups", cell_format = fmt_ae,
               col_header = ae_col_header, col_spec = ae_col_spec,
               col_rel_width = ae_widths, row_height_twips = 200),
  "ae_gtsummary")
```

```{r img-ae-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_gtsummary.png", "AE summary via gtsummary tbl_hierarchical()")
```

## cards + gtsummary

The ARD-first variant: compute the hierarchical **Analysis Results Dataset** with
`cards::ard_stack_hierarchical()`, then summarise it with `tbl_ard_hierarchical()`.
`tbl_ard_hierarchical()` has no `overall_row` argument -- but the any-AE row is
still **the ARD's job, not the raw data's**: a second `ard_stack_hierarchical()`
with **`over_variables = TRUE`** computes the distinct-subject "any event" count
across all the hierarchy variables, and `add_overall_row()` reads it straight off
that ARD (no re-counting of `adae`). Ordering uses the same native
`sort_hierarchical(sort = AE_SORT)`, so the rendered table matches the two above
cell-for-cell.

```{r ae-cards-gtsummary}
library(cards); library(gtsummary)
# Hierarchical ARD for the SOC / PT body.
ard <- ard_stack_hierarchical(
  data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
  denominator = adsl, id = USUBJID)
# "Any event" ARD: over_variables = TRUE adds the distinct-subject overall count.
ard_any <- ard_stack_hierarchical(
  data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
  denominator = adsl, id = USUBJID, over_variables = TRUE)

tbl <- tbl_ard_hierarchical(
    cards = ard, variables = c(AESOC, AEDECOD), by = TRT01A,
    statistic = ~ "{n} ({p}%)") |>
  filter_hierarchical(p >= 0.03) |>
  sort_hierarchical(sort = AE_SORT) |> # SOCs A->Z, PTs by subject count (native)
  add_overall_row(ard_any)             # any-AE row read from the ARD, generically

render_ae(
  as_rtftables(tbl, read_meta = TRUE, split = "group_force", max_rows = AE_MAX_ROWS,
               blank_rows = "between_groups", cell_format = fmt_ae,
               col_header = ae_col_header, col_spec = ae_col_spec,
               col_rel_width = ae_widths, row_height_twips = 200),
  "ae_gtsummary_ard")
```

```{r img-ae-cards-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_gtsummary_ard.png", "AE summary via cards + gtsummary tbl_ard_hierarchical()")
```

> The three AE tables above are the **same report** -- same numbers, same SOC /
> PT order -- built three different ways and all rendered by `rtfreporter`.

## Tplyr (transpose-and-set)

`Tplyr` returns **formatted strings**; as with the DM table we do the
*transpose-and-set* here -- Tplyr counts the distinct subjects, and we assemble
the layout (overall row, SOC count on the SOC row, indented PTs in subject-count
order) and hand the column header to `rtfreporter` via the setup block's
`ae_col_header` / `ae_col_spec`.

```{r ae-tplyr}
library(Tplyr)
varcols <- paste0("var1_", arm_levels)
rn <- function(df) df |> dplyr::rename_with(~ arm_levels, dplyr::all_of(varcols))

# distinct-subject counts: one layer for SOC totals, one for PT-by-SOC.
soc <- tplyr_table(adae, TRT01A) |> set_pop_data(adsl) |> set_pop_treat_var(TRT01A) |>
  add_layer(group_count(AESOC) |> set_distinct_by(USUBJID) |>
              set_format_strings(f_str("xx (xx.x%)", distinct_n, distinct_pct))) |>
  build() |> dplyr::transmute(AESOC = as.character(row_label1),
                              dplyr::across(dplyr::all_of(varcols))) |> rn()
pt <- tplyr_table(adae, TRT01A) |> set_pop_data(adsl) |> set_pop_treat_var(TRT01A) |>
  add_layer(group_count(AEDECOD, by = vars(AESOC)) |> set_distinct_by(USUBJID) |>
              set_format_strings(f_str("xx (xx.x%)", distinct_n, distinct_pct))) |>
  build() |> dplyr::transmute(AESOC = as.character(row_label1),
                              AEDECOD = as.character(row_label2),
                              dplyr::across(dplyr::all_of(varcols))) |>
  dplyr::semi_join(.soc_pt, by = c("AESOC", "AEDECOD")) |>
  dplyr::filter(AEDECOD %in% .keep_pt) |> rn()

# Assemble `disp`: an overall any-AE row, then each SOC row (its count) and its
# PTs indented in subject-count order.
any_n <- adae |> dplyr::distinct(TRT01A, USUBJID) |> dplyr::count(TRT01A)
ov <- vapply(arm_levels, function(a) {
  n <- any_n$n[match(a, as.character(any_n$TRT01A))]
  sprintf("%d (%.1f%%)", n, 100 * n / as.integer(arm_n[a]))
}, character(1))

rows <- list(setNames(data.frame(ANY_AE, t(ov), check.names = FALSE,
                                 stringsAsFactors = FALSE),
                      c("Characteristic", arm_levels)))
for (s in sort(unique(pt$AESOC))) {
  sc <- soc |> dplyr::filter(AESOC == s)
  rows[[length(rows) + 1]] <- setNames(
    data.frame(s, sc[, arm_levels], check.names = FALSE, stringsAsFactors = FALSE),
    c("Characteristic", arm_levels))
  pts <- pt |> dplyr::filter(AESOC == s) |> dplyr::arrange(match(AEDECOD, .pt_order))
  if (nrow(pts))
    rows[[length(rows) + 1]] <- setNames(
      data.frame(paste0("    ", pts$AEDECOD), pts[, arm_levels],
                 check.names = FALSE, stringsAsFactors = FALSE),
      c("Characteristic", arm_levels))
}
disp <- dplyr::bind_rows(rows)

# Metadata lives HERE, in rtfreporter -- not in Tplyr.
render_ae(
  as_rtftables(disp, col_header = ae_col_header, split = "group_force",
               max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
               cell_format = fmt_ae, col_spec = ae_col_spec,
               col_rel_width = ae_widths, row_height_twips = 200),
  "ae_tplyr")
```

```{r img-ae-tplyr, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_tplyr.png", "AE summary via Tplyr (transpose-and-set)")
```

## cards + tfrmt

`tfrmt` is driven by a long **group / label / column / param / value** table. The
trick for an AE table is that a row whose **`group` equals its `label`** renders
as a *single* line that also carries its statistic -- so giving each SOC a
`group = label = <SOC>` row puts the SOC count **on the SOC row**, just like the
others; the PT rows (`group = SOC`, `label = PT`) become the indented children,
and the overall row is its own one-row group. We complete the arm x row grid so
a zero is a real `0`, and sort via `sorting_cols`.

```{r ae-tfrmt}
library(tidyr); library(tfrmt)
N          <- as.integer(arm_n[arm_levels]); names(N) <- arm_levels
real_pairs <- .soc_pt |> dplyr::filter(AEDECOD %in% .keep_pt)
kept_socs  <- sort(unique(real_pairs$AESOC))

# PT counts per arm, completed to a full grid so a missing cell is a real 0.
pt_c <- adae |> dplyr::distinct(TRT01A, USUBJID, AESOC, AEDECOD) |>
  dplyr::count(TRT01A, AESOC, AEDECOD, name = "n") |>
  dplyr::mutate(AESOC = as.character(AESOC), AEDECOD = as.character(AEDECOD)) |>
  dplyr::semi_join(real_pairs, by = c("AESOC", "AEDECOD")) |>
  tidyr::complete(TRT01A, tidyr::nesting(AESOC, AEDECOD), fill = list(n = 0)) |>
  dplyr::mutate(p = n / N[as.character(TRT01A)])
# SOC distinct-subject counts (independent of the PTs).
soc_c <- adae |> dplyr::distinct(TRT01A, USUBJID, AESOC) |>
  dplyr::count(TRT01A, AESOC, name = "n") |>
  dplyr::mutate(AESOC = as.character(AESOC)) |>
  dplyr::filter(AESOC %in% kept_socs) |>
  tidyr::complete(TRT01A, AESOC = kept_socs, fill = list(n = 0)) |>
  dplyr::mutate(p = n / N[as.character(TRT01A)])
# Overall any-AE count.
ov_c <- adae |> dplyr::distinct(TRT01A, USUBJID) |> dplyr::count(TRT01A, name = "n") |>
  dplyr::mutate(p = n / N[as.character(TRT01A)])

# long form: a SOC row has group == label (count shown on the SOC line); PT rows
# are group = SOC, label = PT; the overall row is its own group.  ord1 / ord2 are
# the SOC and PT sort keys.
long <- dplyr::bind_rows(
    ov_c  |> dplyr::transmute(group = ANY_AE, label = ANY_AE,
                              column = as.character(TRT01A), n, p),
    soc_c |> dplyr::transmute(group = AESOC,  label = AESOC,
                              column = as.character(TRT01A), n, p),
    pt_c  |> dplyr::transmute(group = AESOC,  label = AEDECOD,
                              column = as.character(TRT01A), n, p)) |>
  tidyr::pivot_longer(c(n, p), names_to = "param", values_to = "value") |>
  dplyr::mutate(
    ord1   = ifelse(group == ANY_AE, 0L, match(group, kept_socs)),
    ord2   = ifelse(label == group, 0L, match(label, .pt_order)),
    column = factor(column, levels = arm_levels),
    group  = factor(group, levels = c(ANY_AE, kept_socs)))

spec <- tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  sorting_cols = c(ord1, ord2),                       # SOCs A->Z, PTs by count
  body_plan = body_plan(frmt_structure(".default", ".default",
    frmt_combine("{n} ({p}%)", n = frmt("x"),
                 p = frmt("x.x", transform = ~ . * 100)))),
  col_plan = col_plan(group, label, Placebo, `Xanomeline Low Dose`,
                      `Xanomeline High Dose`, -ord1, -ord2))

render_ae(
  as_rtftables(print_to_gt(spec, long), read_meta = TRUE, split = "group_force",
               max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
               cell_format = fmt_ae, col_header = ae_col_header,
               col_spec = ae_col_spec, col_rel_width = ae_widths,
               row_height_twips = 200),
  "ae_tfrmt")
```

```{r img-ae-tfrmt, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_tfrmt.png", "AE summary via cards + tfrmt (group == label puts the count on the SOC row)")
```

## flextable and huxtable (convert and render)

No need to rebuild the table for these two: the gtsummary table converts straight
to a **flextable** or a **huxtable**, and `rtfreporter` reads either. (We bake
the PT indent into the label text first with `bake_indent()`, because the
converters turn gtsummary's indent into cell padding that the flextable /
huxtable adapters do not read as a row-label indent.)

```{r ae-flextable-huxtable}
library(gtsummary); library(flextable); library(huxtable)
# The same gtsummary table as the gtsummary section above.
g <- tbl_hierarchical(
    data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
    denominator = adsl, id = USUBJID, overall_row = TRUE,
    label = list("..ard_hierarchical_overall.." ~ ANY_AE),
    statistic = ~ "{n} ({p}%)", digits = everything() ~ list(p = 1)) |>
  filter_hierarchical(p >= 0.03) |>
  sort_hierarchical(sort = AE_SORT)

# flextable
render_ae(
  as_rtftables(gtsummary::as_flex_table(bake_indent(g)), split = "group_force",
               max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
               cell_format = fmt_ae, col_header = ae_col_header,
               col_spec = ae_col_spec, col_rel_width = ae_widths,
               row_height_twips = 200),
  "ae_flextable")

# huxtable
render_ae(
  as_rtftables(gtsummary::as_hux_table(bake_indent(g)), split = "group_force",
               max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
               cell_format = fmt_ae, col_header = ae_col_header,
               col_spec = ae_col_spec, col_rel_width = ae_widths,
               row_height_twips = 200),
  "ae_huxtable")
```

```{r img-ae-flextable, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_flextable.png", "AE summary via gtsummary -> flextable")
```

```{r img-ae-huxtable, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_huxtable.png", "AE summary via gtsummary -> huxtable")
```

> **Seven ways, one report.** rtables/tern, gtsummary, cards + gtsummary, Tplyr,
> cards + tfrmt, and gtsummary converted to flextable / huxtable -- all the same
> AE table, all rendered to the same paginated RTF by `rtfreporter`.

---

*Coming next, under the same showcase: real Word snapshots in place of the
current placeholders.*
