---
title: "Same report, every table framework"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE, eval = FALSE)

# Show a screenshot of the rendered RTF if it has been captured.  The PNGs sit
# next to the .rtf files in inst/rtf-examples/showcase/; we copy the one we need
# into the article's own figures/ folder (so pkgdown serves it) and include it.
# If the screenshot does not exist yet, a small placeholder is shown instead so
# the article still builds.
.snapshot <- function(file, alt = "rendered RTF") {
  src <- file.path("..", "..", "inst", "rtf-examples", "showcase", file)
  if (file.exists(src)) {
    dir.create(file.path("figures", "showcase"), showWarnings = FALSE,
               recursive = TRUE)
    dest <- file.path("figures", "showcase", file)
    file.copy(src, dest, overwrite = TRUE)
    knitr::include_graphics(dest)
  } else {
    knitr::asis_output(sprintf(
      "<p><em>(snapshot <code>inst/rtf-examples/showcase/%s</code> not captured yet)</em></p>",
      file))
  }
}
```

```{r setup}
library(rtfreporter)
```

The R ecosystem for clinical tables is wonderfully diverse -- **rtables/tern**,
**tfrmt**, **gtsummary** (on its own or on a **cards/cardx** ARD), **Tplyr**,
**gt**, **flextable**, **huxtable** -- and teams pick whichever fits their style
and validation story. What they all have in common is the *last* step: a
**regulatory RTF** that opens cleanly in Word.

That last step is exactly what `rtfreporter` does, for **all of them**. This
article builds the **same demographics report from the same data** with several
frameworks and renders each to RTF through `rtfreporter`. The point is not that
one framework is best -- it is that **whatever you already use, rtfreporter is
the RTF back end**.

> The code below is shown for reference (`eval = FALSE`); each table's RTF is
> rendered offline and the Word output is shown as a screenshot. The full,
> runnable generator lives in
> [`data-raw/showcase_dm.R`](https://github.com/ichirio/rtfreporter/blob/main/data-raw/showcase_dm.R),
> and the rendered `.rtf` files are committed to
> [`inst/rtf-examples/showcase/`](https://github.com/ichirio/rtfreporter/tree/main/inst/rtf-examples/showcase)
> so you can open them yourself.

## The data

We use [`pharmaverseadam`](https://pharmaverse.github.io/pharmaverseadam/)'s
`adsl`, restricted to the three randomised arms, and summarise **age** (as `n`,
`Mean (SD)`, `Median`, `Min, Max`), an **age category**, **sex** and **race** by
treatment -- a production-style Table 14.1 demographics layout.

```{r data}
library(dplyr)
arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl |>
  filter(TRT01A %in% arm_levels) |>
  mutate(
    TRT01A = factor(TRT01A, levels = arm_levels),
    SEX    = factor(SEX, levels = c("M", "F"), labels = c("Male", "Female")),
    AGEGR  = cut(AGE, breaks = c(-Inf, 65, 81, Inf),
                 labels = c("<65", "65 - 80", ">80"), right = FALSE),
    # All race categories as a factor, so even zero-count levels (Asian,
    # American Indian) appear as "0 (0.0%)" in every framework.
    RACE   = factor(RACE, levels = c(
      "WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN",
      "AMERICAN INDIAN OR ALASKA NATIVE"),
      labels = c("White", "Black or African American", "Asian",
                 "American Indian or Alaska Native")))
arm_n <- table(adsl$TRT01A)
```

## The report furniture lives in rtfreporter

The running **header** (sponsor, protocol, page number and the table title) and
the **footer** (analysis note, program path and run time), the **Letter
landscape** page -- all of this is set **once, in rtfreporter**, and is identical
for every framework below. Only the *body* changes from framework to framework.

```{r furniture}
dm_header <- rtf_header(rows = list(
  c(l = "Acme Biopharma, Inc.", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(l = "Protocol ABC-2026-001", r = "Status: Draft"),
  c(c = "Table 14.1.1"),
  c(c = "Summary of Demographic and Baseline Characteristics"),
  c(c = "Safety Population"),
  c(c = "")))
dm_footer <- rtf_footer(rows = list(
  c(l = "Note: Percentages are based on the number of non-missing subjects in each treatment group."),
  c(l = "Program: /prod/abc/tfl/t_14_1_1.R", r = "Generated: 2026-06-17 09:14"),
  c(l = "Source: ADSL", r = "CONFIDENTIAL")))

# Every framework feeds its result to the same call.  `align_count_pct` pads the
# "n (xx.x%)" cells to a common width so they line up in the fixed-width font;
# the generator also gives every page the same column header (with the arm Ns),
# 40 : 20 : 20 : 20 widths, centred data columns, and a blank row between groups
# and at the top / bottom (see data-raw/showcase_dm.R).
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(as_rtftables(tbl, align_count_pct = TRUE))   # `tbl` = framework table
generate_rtfreport(doc, "dm.rtf", overwrite = TRUE)
```

# Demographics (DM)

## Build it once, let rtfreporter read it

These frameworks already carry the column labels, spanning headers, alignment
and cell formatting. `as_rtftables(read_meta = TRUE)` (the default) **reads that
metadata** straight off the object -- you do not re-specify the header.

### gtsummary

```{r gtsummary}
library(gtsummary)
tbl <- adsl |>
  select(TRT01A, AGE, AGEGR, SEX, RACE) |>
  tbl_summary(
    by = TRT01A,
    type = list(AGE ~ "continuous2"),                       # the 4-line age block
    label = list(AGE = "Age (years)", AGEGR = "Age category",
                 SEX = "Sex", RACE = "Race"),
    statistic = list(
      AGE ~ c("{N_nonmiss}", "{mean} ({sd})", "{median}", "{min}, {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(AGE ~ list(N_nonmiss = 0, mean = 1, sd = 2,
                             median = 1, min = 0, max = 0))) |>
  modify_header(all_stat_cols() ~ "**{level}**  \nN = {n}")

as_rtftables(tbl, read_meta = TRUE)   # column labels + Ns read from gtsummary
```

```{r img-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_gtsummary.png", "DM summary via gtsummary")
```

### gtsummary on a cards / cardx ARD

The modern, ARD-first workflow: compute an **Analysis Results Dataset** with
`cards`, summarise it with `gtsummary::tbl_ard_summary()`, then hand the table
to `rtfreporter` exactly as above.

```{r gtsummary-ard}
library(cards); library(gtsummary)
ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE,
                 statistic = ~ list(N = length, mean = mean, sd = sd,
                                    median = median, min = min, max = max)),
  ard_categorical(variables = c(AGEGR, SEX, RACE)), .total_n = TRUE)

tbl <- tbl_ard_summary(
  ard, by = TRT01A,
  type = list(AGE = "continuous2"),
  label = list(AGE = "Age (years)", AGEGR = "Age category",
               SEX = "Sex", RACE = "Race"),
  statistic = list(AGE = c("{N}", "{mean} ({sd})", "{median}", "{min}, {max}"),
                   all_categorical() ~ "{n} ({p}%)"))

as_rtftables(tbl, read_meta = TRUE)
```

```{r img-gtsummary-ard, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_gtsummary_ard.png", "DM summary via gtsummary on a cards/cardx ARD")
```

### rtables / tern

`tern::analyze_vars()` produces a `VTableTree`; `rtfreporter` reads its row
labels, column counts and spanning structure via `formatters::matrix_form()`.

```{r rtables}
library(rtables); library(tern)
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze_vars(
    vars = c("AGE", "AGEGR", "SEX", "RACE"),
    var_labels = c(AGE = "Age (years)", AGEGR = "Age category",
                   SEX = "Sex", RACE = "Race"),
    .stats = c("n", "mean_sd", "median", "range", "count_fraction"))
tbl <- build_table(lyt, adsl)

as_rtftables(tbl, read_meta = TRUE)
```

```{r img-rtables, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_rtables.png", "DM summary via rtables/tern")
```

### cards + tfrmt

The *same* cards ARD as above, rendered with **tfrmt** instead of gtsummary --
so the one ARD drives two different formatters. The ARD is reshaped to tfrmt's
`group / label / column / param / value` long form, a `tfrmt` spec formats it,
and `print_to_gt()` produces the `gt` that `rtfreporter` reads.

```{r cards-tfrmt}
library(cards); library(tfrmt)
ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE,
                 statistic = ~ list(n = length, Mean = mean, SD = sd,
                                    Median = median, Min = min, Max = max)),
  ard_categorical(variables = c(AGEGR, SEX, RACE)))

# reshape the ARD to tfrmt's long form: group / label / column / param / value
long <- as.data.frame(ard) |> ...

# Everything except the categorical "n (pct%)" cell is shared; only the
# count/percent formatting -- and who aligns it -- changes between the two
# renders below.
tfrmt_spec <- function(count_pct) tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  body_plan = body_plan(
    frmt_structure(".default", ".default", frmt("xx.x")),
    frmt_structure("Age (years)", "n", frmt("xx")),
    frmt_structure(".default", "Mean (SD)",
                   frmt_combine("{Mean} ({SD})",
                                Mean = frmt("xx.x"), SD = frmt("xx.xx"))),
    frmt_structure(c("Age category", "Sex", "Race"), ".default", count_pct))
```

#### Who formats the `n (xx.x%)` cell?

This is a nice place to see the same table formatted **two ways**, because the
count/percent cell is exactly where formatters differ.

A small but important detail first: write the percent as **`x.x`**, not `xx.x`.
`frmt("xx.x")` right-pads the integer part, so a one-digit percent prints with a
leading space *inside* the parentheses -- `( 3.5%)`. `frmt("x.x")` gives a clean
`(3.5%)`.

**Way 1 -- tfrmt formats and aligns.** The percent uses `x.x`; the count is
padded to `xx`. `rtfreporter` reads the strings and renders them *as-is*
(`align_count_pct` left off):

```{r tfrmt-self}
g1 <- print_to_gt(
  tfrmt_spec(frmt_combine("{n} ({pct}%)", n = frmt("xx"), pct = frmt("x.x"))),
  long)
as_rtftables(g1, read_meta = TRUE)              # tfrmt's own alignment, untouched
```

```{r img-tfrmt, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_tfrmt.png", "DM summary via cards + tfrmt, formatted by tfrmt")
```

**Way 2 -- rtfreporter formats.** tfrmt does the bare minimum -- just clean
`n (x.x%)` strings, no count padding -- and `rtfreporter` re-aligns them with
`align_count_pct = TRUE`:

```{r tfrmt-rtfreporter}
g2 <- print_to_gt(
  tfrmt_spec(frmt_combine("{n} ({pct}%)", n = frmt("x"), pct = frmt("x.x"))),
  long)
as_rtftables(g2, read_meta = TRUE, align_count_pct = TRUE)   # rtfreporter aligns
```

```{r img-tfrmt-rtfreporter, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_tfrmt_rtfreporter.png", "DM summary via cards + tfrmt, formatted by rtfreporter")
```

The two are deliberately *not* byte-identical. tfrmt anchors the `(` right after
the count, so the percentages stay left-aligned within the parentheses.
`rtfreporter`'s `align_count_pct` instead right-aligns the whole parenthetical so
the **decimal points line up**, and drops the redundant decimal at 100%
(`100%`, not `100.0%`). Use whichever convention your SAP calls for -- the point
is that you can hand `rtfreporter` already-formatted strings *or* let it own the
count/percent alignment, from the very same tfrmt table.

## Shape the body, set the look in rtfreporter

Some frameworks return just the **formatted body** -- the numbers, as strings --
and leave presentation to you. Here you do the opposite of the section above:
the framework transposes/summarises, and **`rtfreporter` supplies the column
header, alignment and grouping**.

### Tplyr

`Tplyr::build()` returns a data frame of formatted strings (plus ordering
helpers). We reshape it into a labelled body and pass the column header to
`rtftable()` ourselves.

```{r tplyr}
library(Tplyr)
built <- tplyr_table(adsl, TRT01A) |>
  add_layer(group_desc(AGE, by = "Age (years)") |>
              set_format_strings("n"         = f_str("xx", n),
                                 "Mean (SD)" = f_str("xx.x (xx.xx)", mean, sd),
                                 "Median"    = f_str("xx.x", median),
                                 "Min, Max"  = f_str("xx, xx", min, max))) |>
  # "x.x" (not "xx.x") so a one-digit percent has no leading space inside the
  # parens -- a clean " 1 (1.2%)" that align_count_pct can re-align.
  add_layer(group_count(AGEGR, by = "Age category") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  add_layer(group_count(SEX, by = "Sex") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  add_layer(group_count(RACE, by = "Race") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  build()

# ... reshape `built` into `disp`: a label column (group headers + indented
#     stats, a "Number of Subjects" row on top) and one column per arm ...

# Metadata lives HERE, in rtfreporter -- not in Tplyr:
col_header <- c("Characteristic", paste0(arm_levels, "\nN = ", arm_n))
rtftable(disp, col_header = col_header)
```

```{r img-tplyr, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("dm_tplyr.png", "DM summary via Tplyr")
```

## Two ways to divide the work

The two sections above are the key choice when wiring any framework to
`rtfreporter`:

| | **Build-and-read** | **Transpose-and-set** |
|---|---|---|
| Who owns the column header / spanning / formatting? | the table framework | `rtfreporter` |
| How rtfreporter gets it | `as_rtftables(read_meta = TRUE)` | you pass `col_header`, `col_spec`, ... |
| Best for | metadata-rich frameworks (rtables/tern, tfrmt, gtsummary) | body-only frameworks (Tplyr, tidytlg) |
| Trade-off | least code; depends on the framework's metadata | full control; you re-state the look once |

Both produce the **same RTF**. Pick whichever keeps the work where it already
lives: if your framework already knows the headers, let `rtfreporter` read them;
if it only knows the numbers, let `rtfreporter` dress them.

# Adverse Events (AE)

The demographics table fits on a page; a real **adverse-events** table does not.
This is where `rtfreporter`'s **pagination** earns its keep: a long
System-Organ-Class / Preferred-Term table is split across pages, and a class
that straddles a page boundary has its label **repeated with `(Cont.)`** at the
top of the next page.

We count the **number of subjects** with at least one treatment-emergent AE,
by treatment group, shown as `n (xx.x%)`:

- **"Subjects with any adverse event"** overall on top, then per **System Organ
  Class (SOC)**, and -- indented -- per **Preferred Term (PT)**.
- Each level is an **independent distinct-subject count**: the SOC count sits on
  the **SOC row itself** and counts a subject once even if they had several PTs
  in that class, so a SOC count is **not** the sum of its PTs.
- Limited to PTs occurring in **>= 2.5% of subjects in any group** (the overall /
  SOC counts are still over *all* TEAEs), so the table is a readable couple of
  pages.
- Rows ordered by **subject count, descending**; ties broken **alphabetically**.

## rtables / tern

`tern` is the pharmaverse standard for this table. `analyze_num_patients()`
gives the overall "any AE" row; a SOC-level content summary
(`summarize_row_groups()` with `child_labels = "hidden"`) puts each SOC's
distinct-subject count **on the SOC name row**; `count_occurrences()` gives the
indented PT rows. We build on the full data, **prune** PT rows below 2.5% with
`rtables::prune_table()`, and **sort** by subject count -- alphabetical factor
levels plus a stable `sort_at_path()` break ties A -> Z. `rtfreporter` then
paginates with **`split = "group_force"`** (`max_rows` per page), which is what
inserts the `(Cont.)` rows.

```{r ae-tern}
library(rtables); library(tern)
adae <- pharmaverseadam::adae |>
  filter(TRT01A %in% arm_levels, TRTEMFL == "Y") |>
  mutate(TRT01A  = factor(TRT01A, levels = arm_levels),
         AESOC   = factor(AESOC,   levels = sort(unique(AESOC))),     # ties A->Z
         AEDECOD = factor(AEDECOD, levels = sort(unique(AEDECOD))))

# distinct subjects with any AE in a SOC, per column -- shown on the SOC row
soc_count <- function(df, labelstr, .N_col, ...) {
  n <- length(unique(df$USUBJID))
  in_rows(rcell(c(n, n / .N_col), format = "xx (xx.x%)"), .labels = labelstr)
}

ae <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze_num_patients(vars = "USUBJID", .stats = "unique",
                       .labels = c(unique = "Subjects with any adverse event")) |>
  split_rows_by("AESOC", child_labels = "hidden", nested = FALSE,
                split_fun = drop_split_levels) |>
  summarize_row_groups(cfun = soc_count) |>              # SOC count on the SOC row
  count_occurrences(vars = "AEDECOD", .indent_mods = 1L) |>
  build_table(df = adae, alt_counts_df = adsl) |>
  prune_table(keep_rows(has_fraction_in_any_col(atleast = 0.025))) |>   # >= 2.5%
  sort_at_path(path = c("AESOC"),                 scorefun = cont_n_allcols) |>
  sort_at_path(path = c("AESOC", "*", "AEDECOD"), scorefun = score_occurrences)

# group_force = fill each page to max_rows and repeat a split SOC with "(Cont.)"
as_rtftables(ae, split = "group_force", max_rows = 32,
             blank_rows = "between_groups", cell_format = fmt_count_paren_bare)
```

```{r img-ae-tern, eval = TRUE, echo = FALSE, out.width = "100%"}
.snapshot("ae_tern.png", "AE summary via rtables/tern, paginated with (Cont.)")
```

The same `as_rtftables()` call returns a **list of pages**; `rtf_tables()`
renders them in order, and the running header's `{AUTO_PAGE}` /
`{AUTO_TOTAL_PAGES}` number them.

---

*Coming next, under the same showcase: the AE table from more frameworks
(**cards + tfrmt**, **Tplyr**, **gtsummary `tbl_hierarchical`**), then real Word
snapshots, plus flextable / huxtable variants.*
