---
title: "From pharmaverse tables to RTF reports"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
.pkgs <- c("tern", "rtables", "gtsummary", "tfrmt", "cards", "forcats",
           "pharmaverseadam", "dplyr")
.ok <- all(vapply(.pkgs, requireNamespace, logical(1L), quietly = TRUE))
knitr::opts_chunk$set(eval = .ok)
```

```{r, eval = TRUE, echo = FALSE, results = "asis"}
if (!.ok)
  cat("> This article renders only when the pharmaverse table packages",
      "(tern, rtables, gtsummary, tfrmt, cards, forcats) and pharmaverseadam",
      "are installed.  The code below shows the workflow either way.\n")
```

## rtfreporter complements the pharmaverse

The [pharmaverse](https://pharmaverse.org) already produces beautiful clinical
tables: [tern](https://insightsengineering.github.io/tern/) /
[rtables](https://insightsengineering.github.io/rtables/),
[gtsummary](https://www.danieldsjoberg.com/gtsummary/) and
[tfrmt](https://gsk-biostatistics.github.io/tfrmt/) each compute the
statistics and format the numbers.  rtfreporter does **not** reinvent any of
that.  It handles the *last mile*: taking whatever table object those packages
produced and laying it out as a submission-style **RTF** page -- with the
pagination, blank rows and running headers a real deliverable needs.

To make that concrete, this article is written as a **continuation of the
official [pharmaverse examples](https://pharmaverse.github.io/examples/)**.
We take the table objects built -- *verbatim* -- by the
[demographic](https://pharmaverse.github.io/examples/tlg/demographic.html) and
[adverse events](https://pharmaverse.github.io/examples/tlg/adverse_events.html)
examples, and add one final step: render each of them to RTF with
[`as_rtftables()`](../reference/as_rtftables.html) +
[`rtf_tables()`](../reference/rtf_tables.html) +
[`generate_rtfreport()`](../reference/generate_rtfreport.html).  The aim is to
show that **whichever pharmaverse package you already use, rtfreporter can
read its object** and finish the job.

> All credit for the tables below belongs to the pharmaverse examples and
> their authors; rtfreporter only adds the RTF rendering step.

# Rendering the pharmaverse example tables

```{r setup, eval = TRUE}
library(rtfreporter)
```

```{r libs}
library(tern)
library(rtables)
library(gtsummary)
library(tfrmt)
library(cards)
library(forcats)
library(dplyr)
```

### Data

Exactly as in the pharmaverse examples, we use the `pharmaverseadam` ADaM
data.

```{r data}
adsl <- pharmaverseadam::adsl
adae <- pharmaverseadam::adae
```

A small reusable clinical header / footer, used for every table below (this is
the only rtfreporter-specific glue; see *Using rtfreporter* for the details):

```{r hf}
make_header <- function(table_no, title2) {
  rtf_header(rows = list(
    c(l = "Hoge Co. Limited",   r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no, "  ", title2))
  ))
}
make_footer <- function(built_with) {
  rtf_footer(c(l = paste0(
    "Table object built with ", built_with,
    " (pharmaverse example); rendered to RTF by rtfreporter.")))
}
```

## Demographics -- three pharmaverse packages

The pharmaverse *demographic* example builds the same table three different
ways.  Each produces a different object class; `as_rtftables()` reads all of
them and the RTF comes out the same clinical shape.

### tern + rtables (a `TableTree`)

```{r dm-tern}
adsl2 <- adsl |>
  df_explicit_na()

vars <- c("AGE", "AGEGR1", "SEX", "RACE")
var_labels <- c("Age (yr)", "Age group", "Sex", "Race")

lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ACTARM") |>
  add_overall_col("All Patients") |>
  analyze_vars(vars = vars, var_labels = var_labels)
dm_tern <- build_table(lyt, adsl2)
dm_tern
```

```{r dm-tern-rtf}
demog_title <- list(c("Demographic and Baseline Characteristics",
                      "Safety Analysis Set"))
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1a", "Demographics (tern)"),
                             footer = make_footer("tern + rtables"))) |>
  rtf_tables(as_rtftables(dm_tern, blank_rows = "between_groups"),
             titles = demog_title)
out_dm_tern <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_dm_tern, overwrite = TRUE)
file.exists(out_dm_tern)
```

### gtsummary + cards (a `gt_tbl`)

```{r dm-gts}
theme_gtsummary_compact()

ard <- ard_stack(
  adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = c(AGEGR1, SEX, RACE)),
  .by = ACTARM,
  .attributes = TRUE)

dm_gts <- tbl_ard_summary(
  cards = ard, by = ACTARM,
  include = c(AGE, AGEGR1, SEX, RACE),
  type = AGE ~ "continuous2",
  statistic = AGE ~ c("{N}", "{mean} ({sd})",
                      "{median} ({p25}, {p75})", "{min}, {max}")) |>
  bold_labels() |>
  modify_header(all_stat_cols() ~ "**{level}**  \nN = {n}") |>
  modify_footnote(everything() ~ NA)
dm_gts
```

```{r dm-gts-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1b", "Demographics (gtsummary)"),
                             footer = make_footer("gtsummary + cards"))) |>
  rtf_tables(as_rtftables(dm_gts, blank_rows = "between_groups"),
             titles = demog_title)
out_dm_gts <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_dm_gts, overwrite = TRUE)
file.exists(out_dm_gts)
```

### tfrmt + cards (a `gt_tbl`)

```{r dm-tfrmt}
ard <- ard_stack(
  adsl,
  ard_continuous(
    variables = AGE,
    statistic = ~ continuous_summary_fns(c("N", "mean", "sd", "min", "max"))),
  ard_categorical(variables = c(AGEGR1, SEX, RACE)),
  .by = ACTARM, .overall = TRUE, .total_n = TRUE)

ard_tbl <- ard |>
  shuffle_card(fill_overall = "Total") |>
  prep_big_n(vars = "ACTARM") |>
  prep_combine_vars(vars = c("AGE", "AGEGR1", "SEX", "RACE")) |>
  prep_label() |>
  group_by(ACTARM, stat_variable) |>
  mutate(across(c(variable_level, label),
                ~ ifelse(stat_name == "N", "n", .x))) |>
  ungroup() |> unique() |>
  mutate(
    ord1 = fct_inorder(stat_variable) |> fct_relevel("SEX", after = 0) |> as.numeric(),
    ord2 = ifelse(label == "n", 1, 2)) |>
  mutate(stat_variable = case_when(
    stat_variable == "AGE"    ~ "Age (YEARS) at First Dose",
    stat_variable == "AGEGR1" ~ "Age Group (YEARS) at First Dose",
    stat_variable == "SEX"    ~ "Sex",
    stat_variable == "RACE"   ~ "High Level Race",
    .default = stat_variable)) |>
  select(ACTARM, stat_variable, label, stat_name, stat, ord1, ord2) |>
  unique()

dm_tfrmt <- tfrmt(
  group = stat_variable, label = label, param = stat_name,
  value = stat, column = ACTARM, sorting_cols = c(ord1, ord2),
  body_plan = body_plan(
    frmt_structure(".default", ".default", frmt("xxx")),
    frmt_structure(".default", ".default",
      frmt_combine("{n} ({p}%)", n = frmt("xxx"),
                   p = frmt("xx", transform = ~ . * 100)))),
  big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)")),
  col_plan = col_plan(-starts_with("ord")),
  col_style_plan = col_style_plan(col_style_structure(
    col = c("Placebo", "Xanomeline High Dose", "Xanomeline Low Dose", "Total"),
    align = "left")),
  row_grp_plan = row_grp_plan(row_grp_structure(
    ".default", element_block(post_space = " ")))) |>
  print_to_gt(ard_tbl)
dm_tfrmt
```

```{r dm-tfrmt-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1c", "Demographics (tfrmt)"),
                             footer = make_footer("tfrmt + cards"))) |>
  rtf_tables(as_rtftables(dm_tfrmt, blank_rows = "between_groups"),
             titles = demog_title)
out_dm_tfrmt <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_dm_tfrmt, overwrite = TRUE)
file.exists(out_dm_tfrmt)
```

## Adverse events -- paginated across many pages

The pharmaverse *adverse events* example is a genuinely large table -- 23
system-organ classes and 242 preferred terms -- so it spans **many** pages.
This is exactly where rtfreporter earns its keep: `as_rtftables()` paginates
the object, repeats the column headers and the running header on every page,
and keeps each SOC block intact.  Again the example builds it two ways.

```{r ae-prep}
adsl_ae <- adsl |>
  df_explicit_na()
adae_ae <- adae |>
  df_explicit_na() |>
  var_relabel(AEBODSYS = "MedDRA System Organ Class",
              AEDECOD  = "MedDRA Preferred Term") |>
  filter(SAFFL == "Y")

ae_title <- list(c("Adverse Events by System Organ Class and Preferred Term",
                   "Safety Analysis Set"))
```

### tern + rtables (a `TableTree`)

```{r ae-tern}
split_fun <- drop_split_levels
ae_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ACTARM") |>
  add_overall_col(label = "All Patients") |>
  analyze_num_patients(
    vars = "USUBJID", .stats = c("unique", "nonunique"),
    .labels = c(
      unique    = "Total number of patients with at least one adverse event",
      nonunique = "Overall total number of events")) |>
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE,
                split_fun = split_fun, label_pos = "topleft",
                split_label = obj_label(adae_ae$AEBODSYS)) |>
  summarize_num_patients(
    var = "USUBJID", .stats = c("unique", "nonunique"),
    .labels = c(
      unique    = "Total number of patients with at least one adverse event",
      nonunique = "Total number of events")) |>
  count_occurrences(vars = "AEDECOD", .indent_mods = -1L) |>
  append_varlabels(adae_ae, "AEDECOD", indent = 1L)
ae_tern <- build_table(ae_lyt, df = adae_ae, alt_counts_df = adsl_ae)

# split = "group_safe" never breaks an SOC across a page boundary.
ae_tern_pages <- as_rtftables(ae_tern, split = "group_safe", max_rows = 30,
                              blank_rows = "between_groups", align_count_pct = TRUE)
length(ae_tern_pages)            # number of RTF pages
```

```{r ae-tern-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.3.1a", "Adverse Events (tern)"),
                             footer = make_footer("tern + rtables"))) |>
  rtf_tables(ae_tern_pages, titles = rep(ae_title, length(ae_tern_pages)))
out_ae_tern <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_ae_tern, overwrite = TRUE)
file.exists(out_ae_tern)
```

### tfrmt + cards (a `gt_tbl`)

```{r ae-tfrmt}
ae_ard <- ard_stack_hierarchical(
  data = adae_ae, by = ACTARM, variables = c(AEBODSYS, AEDECOD),
  statistic = ~ c("n", "p"), denominator = adsl_ae, id = USUBJID,
  over_variables = TRUE, overall = TRUE)
ae_tot <- ard_stack_hierarchical(
  data = mutate(adae_ae, ACTARM = "All Patients"), by = ACTARM,
  variables = c(AEBODSYS, AEDECOD),
  denominator = mutate(adsl_ae, ACTARM = "All Patients"),
  statistic = ~ c("n", "p"), id = USUBJID,
  over_variables = TRUE, overall = TRUE) |>
  filter(group2 == "ACTARM" | variable == "ACTARM")
ae_card <- bind_ard(ae_ard, ae_tot) |>
  shuffle_card(fill_hierarchical_overall = "ANY EVENT") |>
  prep_big_n(vars = "ACTARM") |>
  prep_hierarchical_fill(vars = c("AEBODSYS", "AEDECOD"), fill = "ANY EVENT") |>
  mutate(ACTARM = ifelse(ACTARM == "Overall ACTARM", "All Patients", ACTARM))
ord_soc <- ae_card |>
  filter(ACTARM == "All Patients", stat_name == "n", AEDECOD == "ANY EVENT") |>
  arrange(desc(stat)) |> mutate(ord1 = row_number()) |> select(AEBODSYS, ord1)
ord_pt <- ae_card |>
  filter(ACTARM == "All Patients", stat_name == "n") |>
  group_by(AEBODSYS) |> arrange(desc(stat)) |>
  mutate(ord2 = row_number()) |> select(AEBODSYS, AEDECOD, ord2)
ae_card <- ae_card |>
  full_join(ord_soc, by = "AEBODSYS") |>
  full_join(ord_pt, by = c("AEBODSYS", "AEDECOD")) |>
  select(AEBODSYS, AEDECOD, ord1, ord2, stat, stat_name, ACTARM)

ae_tfrmt <- tfrmt_n_pct(
  n = "n", pct = "p",
  pct_frmt_when = frmt_when(
    "==1" ~ frmt("(100%)"), ">=0.995" ~ frmt("(>99%)"), "==0" ~ frmt(""),
    "<=0.01" ~ frmt("(<1%)"), "TRUE" ~ frmt("(xx.x%)", transform = ~ . * 100))) |>
  tfrmt(
    group = AEBODSYS, label = AEDECOD, param = stat_name, value = stat,
    column = ACTARM, sorting_cols = c(ord1, ord2),
    col_plan = col_plan("System Organ Class / Preferred Term" = AEBODSYS,
                        Placebo, `Xanomeline High Dose`, `Xanomeline Low Dose`,
                        -ord1, -ord2),
    row_grp_plan = row_grp_plan(row_grp_structure(
      group_val = ".default", element_block(post_space = " "))),
    big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)"))) |>
  print_to_gt(ae_card)

# split = "group_force" cuts at max_rows, adding a "(Cont.)" marker.
ae_tfrmt_pages <- as_rtftables(ae_tfrmt, split = "group_force", max_rows = 30,
                               blank_rows = "between_groups")
length(ae_tfrmt_pages)
```

```{r ae-tfrmt-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.3.1b", "Adverse Events (tfrmt)"),
                             footer = make_footer("tfrmt + cards"))) |>
  rtf_tables(ae_tfrmt_pages, titles = rep(ae_title, length(ae_tfrmt_pages)))
out_ae_tfrmt <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_ae_tfrmt, overwrite = TRUE)
file.exists(out_ae_tfrmt)
```

## Assembling a deliverable

A real submission bundles many TLGs into one file.  Render each table to its
own `.rtf`, then concatenate them with
[`assemble_rtf()`](../reference/assemble_rtf.html) (page numbering and section
headers continue across the seam).  Here we bundle the gtsummary demographics
with the full multi-page tern adverse-events table:

```{r assemble}
bundle <- tempfile(fileext = ".rtf")
assemble_rtf(c(out_dm_gts, out_ae_tern), bundle)
file.exists(bundle)
```

# Using rtfreporter

The examples above kept the rtfreporter-specific code to a minimum.  This
section collects the pieces you can tune.

## Headers and footers

`rtf_header()` / `rtf_footer()` take rows of left / centre / right cells
(`l` / `c` / `r`).  The tokens `{AUTO_PAGE}` and `{AUTO_TOTAL_PAGES}` expand to
the current page and the total page count of the section, so a one-line
"Page X of Y" works across a paginated table automatically (as in `make_header`
above).

## Pagination

`as_rtftables()` turns one table object into a list of one-page `rtftable`s.
The `split` argument controls how:

* `"group_safe"` -- never breaks a row group; an SOC that would overflow moves
  whole to the next page.
* `"group_force"` -- cuts strictly at `max_rows`, inserting a `(Cont.)` row so
  the reader knows the group continues.
* `"by_value"` -- one group per page.
* `"rows"` -- split at explicit row positions (`split_rows`).
* `"none"` -- a single page.

`blank_rows = "between_groups"` inserts a blank line between row groups, and
`align_count_pct = TRUE` re-pads `n (xx.x%)` cells so the percentages line up
in a monospaced renderer.

## What carries across, by source

What does and does not survive the conversion (column headers, spanning,
alignment, titles, footnotes) depends on the source object and is documented
in the *What is carried, by source* table in
[`?as_rtftables`](../reference/as_rtftables.html).

## Notes

* The output is `.rtf`; open it in Word / LibreOffice (or batch-convert to
  PDF).
* The pharmaverse example tables are reproduced here unchanged; only the final
  RTF-rendering step is rtfreporter's.
