---
title: "TLG cookbook: pharmaverse tables -> RTF"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
.pkgs <- c("gtsummary", "rtables", "tern", "tfrmt", "random.cdisc.data",
           "pharmaverseadam", "cards", "dplyr", "tidyr")
.ok <- all(vapply(.pkgs, requireNamespace, logical(1L), quietly = TRUE))
knitr::opts_chunk$set(eval = .ok)
```

```{r, eval = TRUE, echo = FALSE, results = "asis"}
if (!.ok)
  cat("> This article renders only when gtsummary, rtables, tern, tfrmt,",
      "random.cdisc.data, dplyr and tidyr are installed.  The code below",
      "shows the workflow either way.\n")
```

rtfreporter does not compute statistics or format numbers -- the
[pharmaverse](https://pharmaverse.org) ecosystem already does that
beautifully.  rtfreporter handles the **last mile**: taking whatever table
object your team produced and laying it out as a clinical-style RTF page,
with the pagination, blank rows and running headers a real deliverable
needs.

This cookbook follows the
[pharmaverse TLG examples](https://pharmaverse.github.io/examples/tlg/):
the *Demographic* table is built **three ways** (tern, gtsummary, tfrmt)
and the *Adverse Events* table is paginated across several pages.  Each is
converted to RTF with [`as_rtftables()`](../reference/as_rtftables.html) +
[`rtf_tables()`](../reference/rtf_tables.html) +
[`generate_rtfreport()`](../reference/generate_rtfreport.html).

```{r setup, eval = TRUE}
library(rtfreporter)
```

```{r libs}
library(rtables)
library(tern)
library(gtsummary)
library(tfrmt)
library(cards)
library(dplyr)
library(tidyr)
```

## Data

For the demographics we use the CDISC pilot ADaM data shipped with
`random.cdisc.data` (the same data the NEST TLG catalog uses).  For the
adverse-events table we use the `pharmaverseadam` ADaM data, exactly as in
the [pharmaverse adverse-events
example](https://pharmaverse.github.io/examples/tlg/adverse_events.html), so
the table objects here are *the very ones produced by that example*.

```{r data}
adsl <- random.cdisc.data::cadsl   # demographics

ae_adsl <- pharmaverseadam::adsl |>
  df_explicit_na()
ae_adae <- pharmaverseadam::adae |>
  df_explicit_na() |>
  var_relabel(AEBODSYS = "MedDRA System Organ Class",
              AEDECOD  = "MedDRA Preferred Term") |>
  filter(SAFFL == "Y")
```

A reusable 3-row clinical header (company / protocol / centred table title)
and a one-line provenance footer:

```{r hf}
make_header <- function(table_no, title2) {
  rtf_header(rows = list(
    c(l = "Hoge Co. Limited",   r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no, "  ", title2))
  ))
}
make_footer <- function(source_object, built_with) {
  rtf_footer(c(l = paste0(
    "Source object: ", source_object, " (", built_with, "), ",
    "converted to RTF by rtfreporter.  ",
    "CDISC pilot data (random.cdisc.data); layout after the pharmaverse ",
    "TLG examples.")))
}
demog_title <- list(c("Demographic and Baseline Characteristics",
                      "Safety Analysis Set"))
```

## 1. Demographics, three ways

The same demographic summary, built with three different pharmaverse tools.
Whichever you prefer, `as_rtftables()` reads its object and the RTF comes
out the same clinical shape.

### 1a. tern + rtables

```{r dm-tern}
dm_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  add_overall_col("All Patients") |>
  analyze_vars(
    vars   = c("AGE", "SEX", "RACE"),
    .stats = c("mean_sd", "median", "range", "count_fraction")) |>
  append_topleft("Characteristic")
dm_tern <- build_table(dm_lyt, adsl)
dm_tern
```

```{r dm-tern-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1a", "Demographics (tern)"),
                             footer = make_footer("rtables TableTree", "tern + rtables"))) |>
  rtf_tables(as_rtftables(dm_tern, blank_rows = "between_groups"),
             titles = demog_title)
out_tern <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_tern, overwrite = TRUE)
file.exists(out_tern)
```

### 1b. gtsummary

```{r dm-gts}
dm_gts <- adsl |>
  select(ARM, AGE, SEX, RACE) |>
  tbl_summary(by = ARM,
              label = list(AGE = "Age (years)", SEX = "Sex", RACE = "Race"),
              statistic = list(all_continuous() ~ "{mean} ({sd})")) |>
  modify_header(label = "Characteristic") |>
  add_overall()
dm_gts
```

```{r dm-gts-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1b", "Demographics (gtsummary)"),
                             footer = make_footer("gt_tbl", "gtsummary::tbl_summary()"))) |>
  rtf_tables(as_rtftables(dm_gts, align_count_pct = TRUE,
                          blank_rows = "between_groups"),
             titles = demog_title)
out_gts <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_gts, overwrite = TRUE)
file.exists(out_gts)
```

### 1c. tfrmt

`tfrmt` describes the table format declaratively (here: `xx.x` for the age
rows, a combined `n (xx.x%)` for the categorical rows) and renders to gt.

```{r dm-tfrmt}
age_long <- adsl |>
  group_by(ARM) |>
  summarise(Mean = mean(AGE), SD = sd(AGE), .groups = "drop") |>
  pivot_longer(c(Mean, SD), names_to = "label", values_to = "value") |>
  mutate(group = "Age (years)", param = tolower(label))

cat_long <- function(var, group_lbl) {
  adsl |>
    count(ARM, !!sym(var)) |>
    group_by(ARM) |>
    mutate(pct = n / sum(n) * 100) |>
    ungroup() |>
    pivot_longer(c(n, pct), names_to = "param", values_to = "value") |>
    mutate(group = group_lbl, label = as.character(.data[[var]])) |>
    select(ARM, group, label, param, value)
}
dm_long <- bind_rows(age_long |> select(ARM, group, label, param, value),
                     cat_long("SEX", "Sex"),
                     cat_long("RACE", "Race"))

tf <- tfrmt(
  group = group, label = label, column = ARM, param = param, value = value,
  body_plan = body_plan(
    frmt_structure("Age (years)", ".default", frmt("xx.x")),
    frmt_structure(".default",    ".default",
                   frmt_combine("{n} ({pct}%)",
                                n = frmt("xx"), pct = frmt("xx.x")))))
dm_tfrmt <- print_to_gt(tf, dm_long)
dm_tfrmt
```

```{r dm-tfrmt-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1c", "Demographics (tfrmt)"),
                             footer = make_footer("gt_tbl", "tfrmt"))) |>
  rtf_tables(as_rtftables(dm_tfrmt, blank_rows = "between_groups"),
             titles = demog_title)
out_tfrmt <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_tfrmt, overwrite = TRUE)
file.exists(out_tfrmt)
```

## 2. Adverse events, paginated across many pages

The full "AEs by SOC and Preferred Term" table is genuinely large -- 23
system-organ classes and 242 preferred terms -- so it spans *many* pages.
This is exactly where rtfreporter earns its keep: `as_rtftables()` paginates
the table, repeats the column headers and the running header on every page,
and keeps each SOC block intact.  We build the table object two ways and run
both through the identical RTF pipeline.

`ae_title` is shared by both versions; the header carries the `Page X of Y`
field so the page count is filled in automatically per version.

```{r ae-title}
ae_title <- list(c("Adverse Events by System Organ Class and Preferred Term",
                   "Safety Analysis Set"))
```

### 2a. tern + rtables

This is the adverse-events layout from the pharmaverse example verbatim
(unique/non-unique patient and event counts per SOC, occurrence counts per
preferred term, indented one level).  `split = "group_safe"` never breaks an
SOC block across a page, and `blank_rows = "between_groups"` separates them.

```{r ae-tern-build}
split_fun <- drop_split_levels
ae_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ACTARM") |>
  add_overall_col(label = "All Patients") |>
  analyze_num_patients(
    vars = "USUBJID", .stats = c("unique", "nonunique"),
    .labels = c(
      unique    = "Total number of patients with at least one adverse event",
      nonunique = "Overall total number of events")) |>
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE,
                split_fun = split_fun, label_pos = "topleft",
                split_label = obj_label(ae_adae$AEBODSYS)) |>
  summarize_num_patients(
    var = "USUBJID", .stats = c("unique", "nonunique"),
    .labels = c(
      unique    = "Total number of patients with at least one adverse event",
      nonunique = "Total number of events")) |>
  count_occurrences(vars = "AEDECOD", .indent_mods = -1L) |>
  append_varlabels(ae_adae, "AEDECOD", indent = 1L)
ae_tern <- build_table(ae_lyt, df = ae_adae, alt_counts_df = ae_adsl)

ae_tern_pages <- as_rtftables(ae_tern,
                              split      = "group_safe",
                              max_rows   = 30,
                              blank_rows = "between_groups",
                              align_count_pct = TRUE)
length(ae_tern_pages)            # number of RTF pages
```

```{r ae-tern-rtf}
# One section starting at page 1 covers every page; rtfreporter repeats its
# header/footer (and the "Page X of Y" field) on each page automatically.
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.3.1a",
                                                  "Adverse Events (tern)"),
                             footer = make_footer("rtables TableTree", "tern + rtables"))) |>
  rtf_tables(ae_tern_pages,
             titles = rep(ae_title, length(ae_tern_pages)))
out_ae_tern <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_ae_tern, overwrite = TRUE)
file.exists(out_ae_tern)
```

### 2b. tfrmt + cards

The same example also builds the table with `cards` (an ARD) and `tfrmt`,
SOC/PT sorted by descending frequency.  It renders to gt, so `as_rtftables()`
reads it as a gt table; `split = "group_force"` paginates at `max_rows`,
adding a `(Cont.)` marker when an SOC spills over.

```{r ae-tfrmt-build}
ae_ard <- ard_stack_hierarchical(
  data = ae_adae, by = ACTARM, variables = c(AEBODSYS, AEDECOD),
  statistic = ~ c("n", "p"), denominator = ae_adsl, id = USUBJID,
  over_variables = TRUE, overall = TRUE)

ae_tot <- ard_stack_hierarchical(
  data = mutate(ae_adae, ACTARM = "All Patients"), by = ACTARM,
  variables = c(AEBODSYS, AEDECOD),
  denominator = mutate(ae_adsl, ACTARM = "All Patients"),
  statistic = ~ c("n", "p"), id = USUBJID,
  over_variables = TRUE, overall = TRUE) |>
  filter(group2 == "ACTARM" | variable == "ACTARM")

ae_card <- bind_ard(ae_ard, ae_tot) |>
  shuffle_card(fill_hierarchical_overall = "ANY EVENT") |>
  prep_big_n(vars = "ACTARM") |>
  prep_hierarchical_fill(vars = c("AEBODSYS", "AEDECOD"), fill = "ANY EVENT") |>
  mutate(ACTARM = ifelse(ACTARM == "Overall ACTARM", "All Patients", ACTARM))

ord_soc <- ae_card |>
  filter(ACTARM == "All Patients", stat_name == "n", AEDECOD == "ANY EVENT") |>
  arrange(desc(stat)) |> mutate(ord1 = row_number()) |> select(AEBODSYS, ord1)
ord_pt <- ae_card |>
  filter(ACTARM == "All Patients", stat_name == "n") |>
  group_by(AEBODSYS) |> arrange(desc(stat)) |>
  mutate(ord2 = row_number()) |> select(AEBODSYS, AEDECOD, ord2)
ae_card <- ae_card |>
  full_join(ord_soc, by = "AEBODSYS") |>
  full_join(ord_pt, by = c("AEBODSYS", "AEDECOD")) |>
  select(AEBODSYS, AEDECOD, ord1, ord2, stat, stat_name, ACTARM)

ae_tfrmt <- tfrmt_n_pct(
  n = "n", pct = "p",
  pct_frmt_when = frmt_when(
    "==1"     ~ frmt("(100%)"),
    ">=0.995" ~ frmt("(>99%)"),
    "==0"     ~ frmt(""),
    "<=0.01"  ~ frmt("(<1%)"),
    "TRUE"    ~ frmt("(xx.x%)", transform = ~ . * 100))) |>
  tfrmt(
    group = AEBODSYS, label = AEDECOD, param = stat_name, value = stat,
    column = ACTARM, sorting_cols = c(ord1, ord2),
    col_plan = col_plan("System Organ Class / Preferred Term" = AEBODSYS,
                        Placebo, `Xanomeline High Dose`, `Xanomeline Low Dose`,
                        -ord1, -ord2),
    row_grp_plan = row_grp_plan(row_grp_structure(
      group_val = ".default", element_block(post_space = " "))),
    big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)"))) |>
  print_to_gt(ae_card)

ae_tfrmt_pages <- as_rtftables(ae_tfrmt,
                               split      = "group_force",
                               max_rows   = 30,
                               blank_rows = "between_groups")
length(ae_tfrmt_pages)
```

```{r ae-tfrmt-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.3.1b",
                                                  "Adverse Events (tfrmt)"),
                             footer = make_footer("gt_tbl", "cards + tfrmt"))) |>
  rtf_tables(ae_tfrmt_pages,
             titles = rep(ae_title, length(ae_tfrmt_pages)))
out_ae_tfrmt <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_ae_tfrmt, overwrite = TRUE)
file.exists(out_ae_tfrmt)
```

## 3. Assembling a deliverable

A real submission bundles many TLGs into one file.  Render each table to its
own `.rtf`, then concatenate them with
[`assemble_rtf()`](../reference/assemble_rtf.html) (page numbering and
section headers continue across the seam).  Here we bundle the gtsummary
demographics with the full multi-page tern adverse-events table:

```{r assemble}
bundle <- tempfile(fileext = ".rtf")
assemble_rtf(c(out_gts, out_ae_tern), bundle)
file.exists(bundle)
```

## Notes

* The output is `.rtf`; open it in Word / LibreOffice (or batch-convert to
  PDF).
* `split = "group_force"` cuts strictly at `max_rows` (inserting `(Cont.)`
  rows), while `"group_safe"` never breaks a group; `"by_value"` puts one
  group per page.  See [`as_rtftables()`](../reference/as_rtftables.html).
* What does and does not carry across from each source is documented in the
  *What is carried, by source* table in `?as_rtftables`.
