---
title: "From pharmaverse tables to RTF reports"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
.pkgs <- c("tern", "rtables", "gtsummary", "tfrmt", "cards", "forcats",
           "pharmaverseadam", "dplyr")
.ok <- all(vapply(.pkgs, requireNamespace, logical(1L), quietly = TRUE))
knitr::opts_chunk$set(eval = .ok)

# Show a screenshot of the rendered RTF if it has been captured (PNGs live in
# vignettes/articles/figures/), otherwise leave a small placeholder so the
# article still renders before the screenshots exist.
.snapshot <- function(file, alt = "rendered RTF") {
  path <- file.path("figures", file)
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    knitr::asis_output(sprintf(
      "<p><em>(snapshot <code>figures/%s</code> not captured yet)</em></p>",
      file))
  }
}
```

```{r, eval = TRUE, echo = FALSE, results = "asis"}
if (!.ok)
  cat("> This article renders only when the pharmaverse table packages",
      "(tern, rtables, gtsummary, tfrmt, cards, forcats) and pharmaverseadam",
      "are installed.  The code below shows the workflow either way.\n")
```

## rtfreporter complements the pharmaverse

The [pharmaverse](https://pharmaverse.org) already produces beautiful clinical
tables: [tern](https://insightsengineering.github.io/tern/) /
[rtables](https://insightsengineering.github.io/rtables/),
[gtsummary](https://www.danieldsjoberg.com/gtsummary/) and
[tfrmt](https://gsk-biostatistics.github.io/tfrmt/) each compute the
statistics and format the numbers.  rtfreporter does **not** reinvent any of
that.  It handles the *last mile*: taking whatever table object those packages
produced and laying it out as a submission-style **RTF** page -- with the
pagination, blank rows and running headers a real deliverable needs.

To make that concrete, this article is written as a **continuation of the
official [pharmaverse examples](https://pharmaverse.github.io/examples/)**.
We take the table objects built -- *verbatim* -- by the
[demographic](https://pharmaverse.github.io/examples/tlg/demographic.html) and
[adverse events](https://pharmaverse.github.io/examples/tlg/adverse_events.html)
examples, and add one final step: render each of them to RTF with
[`as_rtftables()`](../reference/as_rtftables.html) +
[`rtf_tables()`](../reference/rtf_tables.html) +
[`generate_rtfreport()`](../reference/generate_rtfreport.html).  The aim is to
show that **whichever pharmaverse package you already use, rtfreporter can
read its object** and finish the job.

> All credit for the tables below belongs to the pharmaverse examples and
> their authors; rtfreporter only adds the RTF rendering step.

**The rendered `.rtf` files produced by this article are committed to the
repository so you can open them yourself:
[`inst/rtf-examples/`](https://github.com/ichirio/rtfreporter/tree/main/inst/rtf-examples).**
The file names mark which pharmaverse example table each one came from
(e.g. `pharmaverse-adverse-events-tfrmt.rtf`).

# Rendering the pharmaverse example tables

For each table the **complete, self-contained recipe is folded** below: expand
it and one copy gives you everything -- package loading, the `pharmaverseadam`
data, the pharmaverse processing, the table object, and the rtfreporter step
that renders it to RTF.  (The recipes deliberately repeat the package/data/
header boilerplate so each one runs on its own.)  Outside the fold we show only
the table object as the example prints it and a snapshot of the resulting RTF.

```{r setup, eval = TRUE}
library(rtfreporter)
```

## Demographics -- three pharmaverse packages

The pharmaverse *demographic* example builds the same table three different
ways -- a tern `TableTree`, a gtsummary `gt_tbl` and a tfrmt `gt_tbl`.
`as_rtftables()` reads all three and the RTF comes out the same clinical shape.

### tern + rtables (a `TableTree`)

<details><summary>Show the full recipe (pharmaverse example + rtfreporter)</summary>
```{r dm-tern, results = "hide"}
library(rtfreporter); library(tern); library(dplyr)
# --- reusable rtfreporter glue (header / footer / blank line) ---
make_header <- function(table_no, title, subtitle = "Safety Analysis Set") {
  rtf_header(rows = list(
    c(l = "HOGESTER Co. Limited1", r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",     r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no)), c(c = title),
    c(c = paste0("<", subtitle, ">"))))
}
make_footer <- function(built_with) rtf_footer(c(l = paste0(
  "Table object built with ", built_with,
  " (pharmaverse example); rendered to RTF by rtfreporter.")))
blank_title <- function(n = 1L) rep(list(""), n)
writable <- as.integer((11 - 2 * 0.6) * 1440)   # landscape Letter writable width

# --- pharmaverse demographic example (tern + rtables) -> table object ---
adsl <- pharmaverseadam::adsl |> df_explicit_na()
vars <- c("AGE", "AGEGR1", "SEX", "RACE")
var_labels <- c("Age (yr)", "Age group", "Sex", "Race")
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ACTARM") |>
  add_overall_col("All Patients") |>
  analyze_vars(vars = vars, var_labels = var_labels)
dm_tern <- build_table(lyt, adsl)

# --- render to RTF with rtfreporter ---
out_dm_tern <- tempfile(fileext = ".rtf")
rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = make_header("14.1.1a", "Demographic and Baseline Characteristics (tern)"),
    footer = make_footer("tern + rtables"))) |>
  rtf_tables(as_rtftables(dm_tern, blank_rows = "between_groups",
                          auto_width = TRUE, table_width_twips = writable),
             titles = blank_title()) |>
  generate_rtfreport(out_dm_tern, overwrite = TRUE)
```
</details>

```{r dm-tern-show, echo = FALSE}
dm_tern
```
```{r dm-tern-snap, echo = FALSE}
.snapshot("pharmaverse-demographic-tern.png")
```

### gtsummary + cards (a `gt_tbl`)

<details><summary>Show the full recipe (pharmaverse example + rtfreporter)</summary>
```{r dm-gts, results = "hide"}
library(rtfreporter); library(gtsummary); library(cards); library(dplyr)
# --- reusable rtfreporter glue (header / footer / blank line) ---
make_header <- function(table_no, title, subtitle = "Safety Analysis Set") {
  rtf_header(rows = list(
    c(l = "HOGESTER Co. Limited1", r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",     r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no)), c(c = title),
    c(c = paste0("<", subtitle, ">"))))
}
make_footer <- function(built_with) rtf_footer(c(l = paste0(
  "Table object built with ", built_with,
  " (pharmaverse example); rendered to RTF by rtfreporter.")))
blank_title <- function(n = 1L) rep(list(""), n)
writable <- as.integer((11 - 2 * 0.6) * 1440)   # landscape Letter writable width

# --- pharmaverse demographic example (gtsummary + cards) -> table object ---
adsl <- pharmaverseadam::adsl
theme_gtsummary_compact()
ard <- ard_stack(
  adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = c(AGEGR1, SEX, RACE)),
  .by = ACTARM, .attributes = TRUE)
dm_gts <- tbl_ard_summary(
  cards = ard, by = ACTARM,
  include = c(AGE, AGEGR1, SEX, RACE),
  type = AGE ~ "continuous2",
  statistic = AGE ~ c("{N}", "{mean} ({sd})",
                      "{median} ({p25}, {p75})", "{min}, {max}")) |>
  bold_labels() |>
  modify_header(all_stat_cols() ~ "**{level}**  \nN = {n}") |>
  modify_footnote(everything() ~ NA)

# --- render to RTF with rtfreporter ---
out_dm_gts <- tempfile(fileext = ".rtf")
rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = make_header("14.1.1b", "Demographic and Baseline Characteristics (gtsummary)"),
    footer = make_footer("gtsummary + cards"))) |>
  rtf_tables(as_rtftables(dm_gts, blank_rows = "between_groups",
                          auto_width = TRUE, table_width_twips = writable),
             titles = blank_title()) |>
  generate_rtfreport(out_dm_gts, overwrite = TRUE)
```
</details>

```{r dm-gts-show, echo = FALSE}
dm_gts
```
```{r dm-gts-snap, echo = FALSE}
.snapshot("pharmaverse-demographic-gtsummary.png")
```

### tfrmt + cards (a `gt_tbl`)

<details><summary>Show the full recipe (pharmaverse example + rtfreporter)</summary>
```{r dm-tfrmt, results = "hide"}
library(rtfreporter); library(tfrmt); library(cards); library(forcats); library(dplyr)
# --- reusable rtfreporter glue (header / footer / blank line) ---
make_header <- function(table_no, title, subtitle = "Safety Analysis Set") {
  rtf_header(rows = list(
    c(l = "HOGESTER Co. Limited1", r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",     r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no)), c(c = title),
    c(c = paste0("<", subtitle, ">"))))
}
make_footer <- function(built_with) rtf_footer(c(l = paste0(
  "Table object built with ", built_with,
  " (pharmaverse example); rendered to RTF by rtfreporter.")))
blank_title <- function(n = 1L) rep(list(""), n)
writable <- as.integer((11 - 2 * 0.6) * 1440)   # landscape Letter writable width

# --- pharmaverse demographic example (tfrmt + cards) -> table object ---
adsl <- pharmaverseadam::adsl
ard <- ard_stack(
  adsl,
  ard_continuous(
    variables = AGE,
    statistic = ~ continuous_summary_fns(c("N", "mean", "sd", "min", "max"))),
  ard_categorical(variables = c(AGEGR1, SEX, RACE)),
  .by = ACTARM, .overall = TRUE, .total_n = TRUE)
ard_tbl <- ard |>
  shuffle_card(fill_overall = "Total") |>
  prep_big_n(vars = "ACTARM") |>
  prep_combine_vars(vars = c("AGE", "AGEGR1", "SEX", "RACE")) |>
  prep_label() |>
  group_by(ACTARM, stat_variable) |>
  mutate(across(c(variable_level, label), ~ ifelse(stat_name == "N", "n", .x))) |>
  ungroup() |> unique() |>
  mutate(
    ord1 = fct_inorder(stat_variable) |> fct_relevel("SEX", after = 0) |> as.numeric(),
    ord2 = ifelse(label == "n", 1, 2)) |>
  mutate(stat_variable = case_when(
    stat_variable == "AGE"    ~ "Age (YEARS) at First Dose",
    stat_variable == "AGEGR1" ~ "Age Group (YEARS) at First Dose",
    stat_variable == "SEX"    ~ "Sex",
    stat_variable == "RACE"   ~ "High Level Race", .default = stat_variable)) |>
  select(ACTARM, stat_variable, label, stat_name, stat, ord1, ord2) |> unique()
dm_tfrmt <- tfrmt(
  group = stat_variable, label = label, param = stat_name,
  value = stat, column = ACTARM, sorting_cols = c(ord1, ord2),
  body_plan = body_plan(
    frmt_structure(".default", ".default", frmt("xxx")),
    frmt_structure(".default", ".default",
      frmt_combine("{n} ({p}%)", n = frmt("xxx"),
                   p = frmt("xx", transform = ~ . * 100)))),
  big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)")),
  col_plan = col_plan(-starts_with("ord")),
  col_style_plan = col_style_plan(col_style_structure(
    col = c("Placebo", "Xanomeline High Dose", "Xanomeline Low Dose", "Total"),
    align = "left")),
  row_grp_plan = row_grp_plan(row_grp_structure(
    ".default", element_block(post_space = " ")))) |>
  print_to_gt(ard_tbl)

# --- render to RTF with rtfreporter ---
out_dm_tfrmt <- tempfile(fileext = ".rtf")
rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = make_header("14.1.1c", "Demographic and Baseline Characteristics (tfrmt)"),
    footer = make_footer("tfrmt + cards"))) |>
  rtf_tables(as_rtftables(dm_tfrmt, auto_width = TRUE, table_width_twips = writable),
             titles = blank_title()) |>
  generate_rtfreport(out_dm_tfrmt, overwrite = TRUE)
```
</details>

```{r dm-tfrmt-show, echo = FALSE}
dm_tfrmt
```
```{r dm-tfrmt-snap, echo = FALSE}
.snapshot("pharmaverse-demographic-tfrmt.png")
```

## Adverse events -- paginated across many pages

The pharmaverse *adverse events* example is a genuinely large table -- 23
system-organ classes and 242 preferred terms -- so it spans **many** pages.
This is where rtfreporter earns its keep: `as_rtftables()` paginates the
object and repeats the column and running headers on every page.  The example
builds it two ways, which also shows the **two pagination strategies**:

* **`split = "group_safe"`** (tern) never breaks an SOC across a page boundary.
* **`split = "group_force"`** (tfrmt) fills each page to `max_rows` and repeats
  the SOC label with `" (Cont.)"` at the top of the next page when a group is
  split.

The objects are large, so under each folded recipe we show only the snapshot.

### tern + rtables (a `TableTree`)

<details><summary>Show the full recipe (pharmaverse example + rtfreporter)</summary>
```{r ae-tern, results = "hide"}
library(rtfreporter); library(tern); library(dplyr)
# --- reusable rtfreporter glue (header / footer / blank line) ---
make_header <- function(table_no, title, subtitle = "Safety Analysis Set") {
  rtf_header(rows = list(
    c(l = "HOGESTER Co. Limited1", r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",     r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no)), c(c = title),
    c(c = paste0("<", subtitle, ">"))))
}
make_footer <- function(built_with) rtf_footer(c(l = paste0(
  "Table object built with ", built_with,
  " (pharmaverse example); rendered to RTF by rtfreporter.")))
blank_title <- function(n = 1L) rep(list(""), n)
writable <- as.integer((11 - 2 * 0.6) * 1440)   # landscape Letter writable width
ae_col_spec <- c(list(list(col = 1L, align = "left")),
                 lapply(2:5, function(j) list(col = j, align = "center")))

# --- pharmaverse adverse-events example (tern + rtables) -> table object ---
adsl_ae <- pharmaverseadam::adsl |> df_explicit_na()
adae_ae <- pharmaverseadam::adae |> df_explicit_na() |>
  var_relabel(AEBODSYS = "MedDRA System Organ Class",
              AEDECOD  = "MedDRA Preferred Term") |>
  filter(SAFFL == "Y")
split_fun <- drop_split_levels
ae_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ACTARM") |>
  add_overall_col(label = "All Patients") |>
  analyze_num_patients(
    vars = "USUBJID", .stats = c("unique", "nonunique"),
    .labels = c(
      unique    = "Total number of patients with at least one adverse event",
      nonunique = "Overall total number of events")) |>
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE,
                split_fun = split_fun, label_pos = "topleft",
                split_label = obj_label(adae_ae$AEBODSYS)) |>
  summarize_num_patients(
    var = "USUBJID", .stats = c("unique", "nonunique"),
    .labels = c(
      unique    = "Total number of patients with at least one adverse event",
      nonunique = "Total number of events")) |>
  count_occurrences(vars = "AEDECOD", .indent_mods = -1L) |>
  append_varlabels(adae_ae, "AEDECOD", indent = 1L)
ae_tern <- build_table(ae_lyt, df = adae_ae, alt_counts_df = adsl_ae)

# --- render to RTF with rtfreporter (wide label col, centred data, fmt_count_paren) ---
ae_tern_pages <- as_rtftables(ae_tern, split = "group_safe", max_rows = 36,
                              blank_rows = "between_groups",
                              cell_format = fmt_count_paren,
                              col_rel_width = c(0.50, 0.125, 0.125, 0.125, 0.125),
                              col_spec = ae_col_spec, row_height_twips = 200)
out_ae_tern <- tempfile(fileext = ".rtf")
rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = make_header("14.3.1a", "Adverse Events by System Organ Class and Preferred Term (tern)"),
    footer = make_footer("tern + rtables"))) |>
  rtf_tables(ae_tern_pages, titles = blank_title(length(ae_tern_pages))) |>
  generate_rtfreport(out_ae_tern, overwrite = TRUE)
```
</details>

```{r ae-tern-snap, echo = FALSE}
.snapshot("pharmaverse-adverse-events-tern.png")
```

### tfrmt + cards (a `gt_tbl`)

<details><summary>Show the full recipe (pharmaverse example + rtfreporter)</summary>
```{r ae-tfrmt, results = "hide"}
library(rtfreporter); library(tfrmt); library(cards); library(dplyr)
# --- reusable rtfreporter glue (header / footer / blank line) ---
make_header <- function(table_no, title, subtitle = "Safety Analysis Set") {
  rtf_header(rows = list(
    c(l = "HOGESTER Co. Limited1", r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",     r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no)), c(c = title),
    c(c = paste0("<", subtitle, ">"))))
}
make_footer <- function(built_with) rtf_footer(c(l = paste0(
  "Table object built with ", built_with,
  " (pharmaverse example); rendered to RTF by rtfreporter.")))
blank_title <- function(n = 1L) rep(list(""), n)
writable <- as.integer((11 - 2 * 0.6) * 1440)   # landscape Letter writable width
ae_col_spec <- c(list(list(col = 1L, align = "left")),
                 lapply(2:5, function(j) list(col = j, align = "center")))

# --- pharmaverse adverse-events example (tfrmt + cards) -> table object ---
adsl_ae <- pharmaverseadam::adsl |> df_explicit_na()
adae_ae <- pharmaverseadam::adae |> df_explicit_na() |>
  var_relabel(AEBODSYS = "MedDRA System Organ Class",
              AEDECOD  = "MedDRA Preferred Term") |>
  filter(SAFFL == "Y")
ae_ard <- ard_stack_hierarchical(
  data = adae_ae, by = ACTARM, variables = c(AEBODSYS, AEDECOD),
  statistic = ~ c("n", "p"), denominator = adsl_ae, id = USUBJID,
  over_variables = TRUE, overall = TRUE)
ae_tot <- ard_stack_hierarchical(
  data = mutate(adae_ae, ACTARM = "All Patients"), by = ACTARM,
  variables = c(AEBODSYS, AEDECOD),
  denominator = mutate(adsl_ae, ACTARM = "All Patients"),
  statistic = ~ c("n", "p"), id = USUBJID,
  over_variables = TRUE, overall = TRUE) |>
  filter(group2 == "ACTARM" | variable == "ACTARM")
ae_card <- bind_ard(ae_ard, ae_tot) |>
  shuffle_card(fill_hierarchical_overall = "ANY EVENT") |>
  prep_big_n(vars = "ACTARM") |>
  prep_hierarchical_fill(vars = c("AEBODSYS", "AEDECOD"), fill = "ANY EVENT") |>
  mutate(ACTARM = ifelse(ACTARM == "Overall ACTARM", "All Patients", ACTARM))
ord_soc <- ae_card |>
  filter(ACTARM == "All Patients", stat_name == "n", AEDECOD == "ANY EVENT") |>
  arrange(desc(stat)) |> mutate(ord1 = row_number()) |> select(AEBODSYS, ord1)
ord_pt <- ae_card |>
  filter(ACTARM == "All Patients", stat_name == "n") |>
  group_by(AEBODSYS) |> arrange(desc(stat)) |>
  mutate(ord2 = row_number()) |> select(AEBODSYS, AEDECOD, ord2)
ae_card <- ae_card |>
  full_join(ord_soc, by = "AEBODSYS") |>
  full_join(ord_pt, by = c("AEBODSYS", "AEDECOD")) |>
  select(AEBODSYS, AEDECOD, ord1, ord2, stat, stat_name, ACTARM)
ae_tfrmt <- tfrmt_n_pct(
  n = "n", pct = "p",
  pct_frmt_when = frmt_when(
    "==1" ~ frmt("(100%)"), ">=0.995" ~ frmt("(>99%)"), "==0" ~ frmt(""),
    "<=0.01" ~ frmt("(<1%)"), "TRUE" ~ frmt("(xx.x%)", transform = ~ . * 100))) |>
  tfrmt(
    group = AEBODSYS, label = AEDECOD, param = stat_name, value = stat,
    column = ACTARM, sorting_cols = c(ord1, ord2),
    col_plan = col_plan("System Organ Class / Preferred Term" = AEBODSYS,
                        Placebo, `Xanomeline High Dose`, `Xanomeline Low Dose`,
                        -ord1, -ord2),
    row_grp_plan = row_grp_plan(row_grp_structure(
      group_val = ".default", element_block(post_space = " "))),
    big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)"))) |>
  print_to_gt(ae_card)

# --- render to RTF with rtfreporter (multi-line headers; group_force pagination) ---
ae_hdr <- c("System Organ Class /\nPreferred Term", "Placebo\n(N=86)",
            "Xanomeline\nHigh Dose\n(N=72)",
            "Xanomeline\nLow Dose\n(N=96)", "All Patients\n(N=306)")
ae_tfrmt_pages <- as_rtftables(ae_tfrmt, split = "group_force", max_rows = 36,
                               col_header = ae_hdr,
                               col_rel_width = c(0.50, 0.125, 0.125, 0.125, 0.125),
                               col_spec = ae_col_spec, cell_format = fmt_count_paren,
                               row_height_twips = 200)
out_ae_tfrmt <- tempfile(fileext = ".rtf")
rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = make_header("14.3.1b", "Adverse Events by System Organ Class and Preferred Term (tfrmt)"),
    footer = make_footer("tfrmt + cards"))) |>
  rtf_tables(ae_tfrmt_pages, titles = blank_title(length(ae_tfrmt_pages))) |>
  generate_rtfreport(out_ae_tfrmt, overwrite = TRUE)
```
</details>

```{r ae-tfrmt-snap, echo = FALSE}
.snapshot("pharmaverse-adverse-events-tfrmt.png")
```

## Assembling a deliverable

A real submission bundles many TLGs into one file.  Render each table to its
own `.rtf`, then concatenate them with
[`assemble_rtf()`](../reference/assemble_rtf.html) (page numbering and section
headers continue across the seam).  Here we bundle the gtsummary demographics
with the full multi-page tern adverse-events table:

```{r assemble}
bundle <- tempfile(fileext = ".rtf")
assemble_rtf(c(out_dm_gts, out_ae_tern), bundle)
file.exists(bundle)
```

# Using rtfreporter

The examples above kept the rtfreporter-specific code to a minimum.  This
section collects the pieces you can tune.

## Headers and footers

`rtf_header()` / `rtf_footer()` take rows of left / centre / right cells
(`l` / `c` / `r`).  The tokens `{AUTO_PAGE}` and `{AUTO_TOTAL_PAGES}` expand to
the current page and the total page count of the section, so a one-line
"Page X of Y" works across a paginated table automatically (as in `make_header`
above).

## Pagination

`as_rtftables()` turns one table object into a list of one-page `rtftable`s.
The `split` argument controls how:

* `"group_safe"` -- never breaks a row group; an SOC that would overflow moves
  whole to the next page.
* `"group_force"` -- cuts strictly at `max_rows`, inserting a `(Cont.)` row so
  the reader knows the group continues.
* `"by_value"` -- one group per page.
* `"rows"` -- split at explicit row positions (`split_rows`).
* `"none"` -- a single page.

`blank_rows = "between_groups"` inserts a blank line between row groups, and
`align_count_pct = TRUE` re-pads `n (xx.x%)` cells so the percentages line up
in a monospaced renderer.

## Cell formatting

Different table packages write count/percent cells in different notations, so
rtfreporter does not hard-code one.  Instead, `as_rtftables(cell_format = )`
takes a **cell-format function** that is applied column-by-column to the body
just before pagination.  rtfreporter ships a few:

* `fmt_count_paren()` -- the workhorse, used by **both** adverse-events tables.
  It scans the whole column, then right-justifies the integer count and
  right-justifies the number *inside* the parentheses, so a column mixing
  `"4 ( 4.7%)"`, `"10 (11.6%)"`, `"3 (<1%)"`, `"70 (100%)"` and a lone `"0"`
  all share one width and line up on both the count digit and the percentage.
  (It also copes with the 4-digit event totals in the tern table, which the
  fixed-width helper below cannot.)
* `align_count_pct = TRUE` / `realign_count_pct()` -- the older fixed-width
  helper for plain `"n (xx.x%)"` cells; kept for back-compatibility.
* `fmt_right_align()` -- the simplest one: right-justify a column to a common
  width.

A cell-format function follows a small **contract**:

> It takes one column -- a character vector -- and returns a character vector
> of the **same length**.  Cells it should not touch are returned unchanged,
> and any padding uses the non-breaking space (U+00A0, written `"\u00a0"`) so
> Word does not collapse it.

When none of the built-ins matches your data, write your own.
`fmt_right_align()` is the template -- here is the whole of it, which you can
copy and adapt:

```{r custom-fmt}
my_right_align <- function(x, nbsp = "\u00a0") {
  x  <- as.character(x)
  nz <- nzchar(trimws(x))                       # the non-empty cells
  w  <- max(nchar(x[nz]), 0L)                   # widest cell in the column
  x[nz] <- formatC(x[nz], width = w, flag = "") # right-justify to that width
  gsub(" ", nbsp, x, fixed = TRUE)              # spaces -> non-breaking spaces
}
my_right_align(c("12 (3.6%)", "0", "1 (0.5%)"))
```

Pass it with `cell_format = my_right_align` (a single function applies to every data
column; a list applies one function per column).

## What carries across, by source

What does and does not survive the conversion (column headers, spanning,
alignment, titles, footnotes) depends on the source object and is documented
in the *What is carried, by source* table in
[`?as_rtftables`](../reference/as_rtftables.html).

## Notes

* The output is `.rtf`; open it in Word / LibreOffice (or batch-convert to
  PDF).
* The pharmaverse example tables are reproduced here unchanged; only the final
  RTF-rendering step is rtfreporter's.
